% !TeX root = ../main.tex
% chapters/discussion.tex

\chapter{Discussion}
\label{ch:discussion}

% ── Comparison with BKV 2009 ──────────────────────────────────────
\section{Comparison with Benton--Kennedy--Varming}
\label{sec:disc-bkv}

This thesis takes the Benton--Kennedy--Varming (BKV)
library~\cite{BKV2009} as its classical foundation
(\cref{ch:introduction}).
BKV formalized $\omega$-CPOs, the Kleene fixed-point theorem, and a
complete denotational semantics for PCF$_v$---including soundness and
computational adequacy---in Coq.
Our library reproduces every construction described in that paper and
extends it with CPO-enriched categories, recursive domain equations,
Scott topology, and the beginnings of a quantum domain theory.
\Cref{tab:bkv-summary} summarizes the major points of divergence;
\cref{app:bkv-migration} provides a comprehensive component-by-component
comparison.

\begin{table}[ht]
\centering
\caption{Summary comparison with BKV~\cite{BKV2009}.}
\label{tab:bkv-summary}
\small
\begin{tabular}{@{}l l l@{}}
\toprule
\textbf{Aspect} & \textbf{BKV paper} & \textbf{This thesis} \\
\midrule
Proof assistant    & Coq (circa 2009)
                   & \Rocq{} 9.0 \\
Structure toolkit  & \texttt{Module}/\texttt{Record}
                   & Hierarchy Builder (\HB{}) \\
CPO base           & Preorder
                   & Partial order \\
Equality           & Setoid \texttt{==}
                   & Leibniz \texttt{=} \\
Lift monad         & Coinductive \texttt{delay}
                   & Flat \texttt{option} + \texttt{ClassicalEpsilon} \\
Supremum proofs    & Described abstractly
                   & Proved constructively (2{,}153 lines) \\
Domain equations   & Described (approx.\ sequences)
                   & Full framework; 1 axiom \\
Enriched categories & Implicit (self-enrichment)
                   & Explicit \HB{} structure \\
PCF adequacy       & Strategy described
                   & Fully proved (820 lines) \\
Quantum extensions & ---
                   & qCPOs following KLM~(2024) \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Infrastructure: Hierarchy Builder.}
The most pervasive difference is infrastructural.
BKV's library predates the Hierarchy Builder framework~\cite{HB2023}
and organizes its hierarchy through hand-written modules, records, and
coercions.
Adding a new layer---such as CPO-enriched categories---would require
manually threading coercions through every dependent definition, a
process that scales poorly.
Our library uses \HB{} to declare the hierarchy
\texttt{HasLe} $\to$ \texttt{Preorder} $\to$ \texttt{PartialOrder}
$\to$ \texttt{CPO} $\to$ \texttt{PointedCPO},
with each mixin contributing exactly the new axioms needed at that level.
\HB{} automatically synthesizes join points, canonical projections, and
sort coercions, so registering a type as a \texttt{CPO.type}
immediately makes it available wherever a \texttt{Preorder.type} is
expected.
The practical consequence is that the enriched-category and
domain-equation layers (\cref{sec:meth-enriched,sec:meth-domain-equations})
were added as extensions of the existing hierarchy rather than
retroactive restructurings.

\paragraph{Equality: Leibniz vs.\ setoid.}
BKV build their hierarchy on preorders rather than partial orders,
so two elements $x$ and $y$ need not be propositionally equal even
when $x \sqsubseteq y$ and $y \sqsubseteq x$.
Their notion of equality is the \emph{kernel} of the preorder---
$x == y \;\text{iff}\; x \sqsubseteq y \land y
\sqsubseteq x$---carried as a setoid throughout the development.
On their coinductive lift, this kernel equality happens to coincide
with weak bisimulation: \texttt{now d} and \texttt{later (now d)} are
mutually below in the convergence order and hence identified by
\texttt{==}, even though they are propositionally distinct.
This is precisely the flexibility that allows the coinductive lift to
function as a CPO in BKV's framework.

Our hierarchy, by contrast, requires antisymmetry at the
\texttt{PartialOrder} level (following Abramsky and
Jung~\cite{AJ1994}), which collapses the kernel equality into
Leibniz~\texttt{=}: if $x \sqsubseteq y$ and $y \sqsubseteq x$, then
$x = y$.
The coinductive lift becomes incompatible under this requirement,
because \Rocq{} can refute
$\texttt{now}\;d = \texttt{later}\;(\texttt{now}\;d)$ by
\texttt{discriminate}, yet the convergence order places them mutually
below each other (\cref{sec:disc-lift-monad}).
We therefore adopt the flat lift $D_\bot = \texttt{option}\;D$
(\cref{sec:meth-lift}), accepting the use of
\texttt{ClassicalEpsilon} in the supremum proof in exchange for
globally Leibniz-compatible types.
The trade-off is discussed further in \cref{sec:disc-tradeoffs}.

\paragraph{The lift monad.}
BKV define a coinductive \texttt{delay D} type whose elements are
either immediate values (\texttt{now d}) or computation steps
(\texttt{later t}).
The bottom element is the infinite loop
$\bot = \texttt{cofix}\;\bot := \texttt{later}\;\bot$, and equality
is weak bisimulation.
Our library uses the flat lift $D_\bot = \texttt{option}\;D$, where
$\bot = \texttt{None}$ and values are $\texttt{Some}\;d$.
The monad laws hold by Leibniz equality and the supremum construction,
while non-trivial, avoids coinductive reasoning entirely.
\Cref{sec:disc-lift-monad} develops the coinductive alternative in
full and identifies the precise obstruction to making \texttt{delay D}
a \texttt{CPO.type} without quotient types.

\paragraph{New contributions.}
Five components of our library have no counterpart in
BKV~\cite{BKV2009}:
\begin{enumerate}
  \item \textbf{CPO-enriched categories.}
    BKV implicitly rely on composition of continuous functions being
    continuous; we formalize this as an explicit \HB{} structure
    (\texttt{CPOEnrichedCat}) with hom-CPOs and jointly continuous
    composition (\cref{sec:meth-enriched}).

  \item \textbf{Enriched natural transformations and Yoneda.}
    Natural transformations between locally continuous functors, with
    pointwise order, vertical composition, and whiskering. The
    enriched Yoneda lemma establishes a CPO isomorphism between
    natural transformations out of a representable functor and the
    representing object (\cref{sec:res-yoneda}).

  \item \textbf{Recursive domain equations.}
    EP-pairs, mixed-variance locally continuous bifunctors, the
    approximation tower, and the bilimit specification.
    The \texttt{ROLL}/\texttt{UNROLL} isomorphism and the deflation-chain
    completeness property are fully derived from a single axiom
    (\texttt{bilimit\_exists}); see \cref{sec:res-domain-equations}.

  \item \textbf{Scott topology.}
    The way-below relation, Scott-open sets, and an equivalence between
    order-theoretic and topological continuity
    (\cref{sec:bg-scott-topology}).

  \item \textbf{Quantum domain theory.}
    Involutive quantales, quantum posets, quantum CPOs, and the quantum
    Kleene fixed-point theorem, following
    Kornell--Lindenhovius--Mislove~\cite{KLM2024}.
    This layer is discussed in \cref{sec:disc-quantum}.
\end{enumerate}

\noindent
Where our library \emph{reproduces} BKV constructions---PCF syntax,
operational and denotational semantics, soundness, adequacy---the
mathematical content is the same but the proofs are carried out on the
modernized \HB{} infrastructure.
The adequacy proof follows BKV's logical-relation structure; our
contribution is not the proof strategy but the end-to-end validation
that the new infrastructure supports a complete language formalization
(\cref{sec:res-soundness,sec:res-adequacy}).

% ── Design trade-offs ─────────────────────────────────────────────
\section{Design Trade-offs}
\label{sec:disc-tradeoffs}

The preceding section compares our library with BKV at the component
level.  This section steps back to assess three global engineering
compromises: the axiom budget, the two-tier packaging architecture, and
the single unproved axiom.

\paragraph{Axiom budget.}
\Cref{tab:axiom-budget} lists every axiom on which the library depends.
The first four are standard in MathComp-adjacent \Rocq{} developments:
\texttt{ClassicalEpsilon} provides the informative excluded middle and
witness extraction needed by the lift supremum
(\cref{sec:meth-lift}); \texttt{Classical} contributes propositional
excluded middle to the Scott-topology closure proof and the adequacy
admissibility argument; \texttt{FunctionalExtensionality} and
\texttt{ProofIrrelevance} are used to establish antisymmetry for
function-space and natural-transformation types, where two elements
that agree pointwise must be propositionally equal.

% \begin{table}[ht]
% \centering
% \caption{Axioms used by the library.}
% \label{tab:axiom-budget}
% \small
% \begin{tabular}{@{}l l l@{}}
% \toprule
% \textbf{Axiom} & \textbf{Role} & \textbf{Scope} \\
% \midrule
% \texttt{ClassicalEpsilon}
%   & Informative excluded middle,
%   & \texttt{Lift.v}, \texttt{Nat.v}, \\
%   & witness extraction
%   & \texttt{PCF\_Denotational.v} \\[3pt]
% \texttt{Classical}
%   & Propositional excluded middle
%   & \texttt{ScottTopology.v}, \\
%   & (\texttt{classic}, \texttt{NNPP})
%   & \texttt{PCF\_Adequacy.v} \\[3pt]
% \texttt{FunctionalExtensionality}
%   & Pointwise function equality
%   & \texttt{CPOTheory.v}, \texttt{FunctionSpaces.v}, \\
%   & & \texttt{NatTrans.v}, \texttt{qCPOProperties.v} \\[3pt]
% \texttt{ProofIrrelevance}
%   & Proof-irrelevant \texttt{Prop}
%   & \texttt{Morphisms.v}, \texttt{FunctionSpaces.v}, \\
%   & & \texttt{NatTrans.v}, \texttt{qCPOProperties.v} \\[3pt]
% \texttt{bilimit\_exists}
%   & Bilimit of approximation sequence
%   & \texttt{DomainEquations.v} \\
% \bottomrule
% \end{tabular}
% \end{table}

% \noindent
The only custom axiom is \texttt{bilimit\_exists}; all others are
shipped with the \Rocq{} standard library.
A consequence of \texttt{ClassicalEpsilon} is that the library's
computational content cannot be extracted to OCaml or Haskell---the
informative excluded middle produces a term that the extraction
mechanism refuses to compile.
For a library whose purpose is denotational \emph{reasoning} rather
than executable \emph{computation}, this is an acceptable cost; no
downstream file uses \texttt{Extraction}, and the axiom is confined
to the lift and Nat supremum constructions.

\begin{table}[ht]
\centering
\caption{Axioms used by the library.}
\label{tab:axiom-budget}
\small
\begin{tabular}{@{}l l l@{}}
\toprule
\textbf{Axiom} & \textbf{Role} & \textbf{Scope} \\
\midrule
\texttt{ClassicalEpsilon}
  & Informative excluded middle,
  & \texttt{Lift.v}, \texttt{Nat.v}, \\
  & witness extraction
  & \texttt{PCF\_Denotational.v} \\[3pt]
\texttt{Classical}
  & Propositional excluded middle
  & \texttt{ScottTopology.v}, \\
  & (\texttt{classic}, \texttt{NNPP})
  & \texttt{PCF\_Adequacy.v} \\[3pt]
\texttt{FunctionalExtensionality}
  & Pointwise function equality
  & \texttt{CPOTheory.v}, \texttt{FunctionSpaces.v}, \\
  & & \texttt{NatTrans.v}, \texttt{qCPOProperties.v} \\[3pt]
\texttt{ProofIrrelevance}
  & Proof-irrelevant \texttt{Prop}
  & \texttt{Morphisms.v}, \texttt{FunctionSpaces.v}, \\
  & & \texttt{NatTrans.v}, \texttt{qCPOProperties.v} \\[3pt]
\texttt{bilimit\_exists}
  & Bilimit of approximation sequence
  & \texttt{DomainEquations.v} \\
\bottomrule
\end{tabular}
\end{table}


\paragraph{The two-tier architecture.}
The library packages its type hierarchy
(\texttt{HasLe} $\to$ \texttt{Preorder} $\to$ $\cdots$ $\to$
\texttt{PointedCPO}) via \HB{}, but uses plain \texttt{Record} types
for morphisms (\texttt{mono\_fun}, \texttt{cont\_fun}), data
(\texttt{chain}), and enriched constructs (\texttt{lc\_functor},
\texttt{nat\_trans}).
The reason is structural: \HB{} structures are indexed by a single
carrier type, whereas morphisms are parameterised by a
domain--codomain pair and functors by a source--target pair.
Attempts to register \texttt{nat\_trans} or \texttt{lc\_functor} as
\HB{} structures trigger universe inconsistencies when instantiated
at the self-enriched category of CPOs
(\cref{sec:meth-nat-trans,sec:meth-lc-functors}; DD-018, DD-019 in
\cref{app:design-decisions}).

This two-tier split is not a temporary workaround.
The single-carrier indexing constraint is fundamental to \HB{}'s
design---it is what allows automatic coercion synthesis along the
hierarchy diamond.
Migrating \texttt{cont\_fun} alone would touch over 50 call sites
for minimal gain, because the plain record already composes well
via the combinator library (\texttt{cont\_comp}, \texttt{cont\_pair},
etc.).
The practical cost is that \texttt{lc\_functor} and
\texttt{nat\_trans} do not participate in \HB{}'s automatic inference,
so the bridge function \texttt{lc\_functor\_of\_hb} is needed when
converting between the two tiers.
A related limitation is that the library has not been audited for
universe polymorphism; the plain-record strategy avoids the issue
entirely at the cost of foreclosing some generic constructions.

\paragraph{The bilimit axiom.}
\texttt{bilimit\_exists} is the library's only custom axiom.
It asserts that every approximation sequence
$D_0 \hookrightarrow F(D_0, D_0) \hookrightarrow \cdots$ has a
bilimit---the enriched-categorical analogue of an inverse limit
(\cref{sec:meth-domain-equations}).
The axiom is structured as a fully specified \texttt{BilimitData}
record: it does not merely assert the existence of a bilimit object
$D_\infty$, but packages the cone ep-pairs, the ROLL/UNROLL
isomorphism, and the deflation-chain identity
$\mathrm{id}_{D_\infty} = \bigsqcup_n (e_n \circ p_n)$.

The proof strategy is well understood: construct the $\omega$-product
CPO $\prod_n D_n$, then extract the coherent sub-CPO of
compatible tuples (Smyth--Plotkin, Abramsky--Jung~\S3.3~\cite{AJ1994}).
The missing piece is the $\omega$-product CPO itself---a countable
product of CPOs equipped with the componentwise order and
componentwise suprema---which the library does not yet provide.
Because all downstream theorems (the ROLL/UNROLL isomorphism, the
deflation chain, the recursive-domain worked example in
\cref{sec:res-domain-equations}) are derived from the
\texttt{BilimitData} record fields alone, completing
\texttt{bilimit\_exists} would not change a single line of any other
file.

\paragraph{Summary.}
The three compromises are orthogonal: the classical axioms concern the
lift and Scott topology, the two-tier architecture concerns the
packaging of morphisms and functors, and the bilimit axiom concerns
domain equations.
Each could be addressed independently in future work---replacing
\texttt{ClassicalEpsilon} with a decidability instance for finite
types, registering \texttt{cont\_fun} as an \HB{} structure if the
single-carrier constraint is relaxed, or constructing the
$\omega$-product CPO---without affecting the other two.

% ── Quantum extensions ────────────────────────────────────────────
\section{Toward Quantum Programming Language Semantics}
\label{sec:disc-quantum}

The classical domain theory developed in Chapters~\ref{ch:methods}
and~\ref{ch:results} models programs whose control flow is
deterministic: given the same inputs, each branch decision produces
the same outcome.
Quantum programming languages challenge this assumption.
This section introduces the physical and mathematical context (\cref{sec:disc-quantum-bg}),
describes our formalisation of quantum CPOs (\cref{sec:disc-qcpo}),
and identifies the structural observation that connects classical and
quantum domain theory via $\CPO$-enrichment
(\cref{sec:disc-quantum-enrichment}).

% ·· Background ··············································
\subsection{Background: Quantum Information and Quantum Order}
\label{sec:disc-quantum-bg}

This subsection develops the physical and mathematical concepts
needed to motivate quantum CPOs, beginning from standard linear
algebra and building to the quantale-valued order structures that
underpin the formalisation in \cref{sec:disc-qcpo}.
The exposition is self-contained; the reader familiar with quantum
information theory may skip to \cref{sec:disc-qcpo}. For those unfamiliar,
we assume some familiarity with linear algebra and analysis for the sake of brevity.

% ── Part 1: Hilbert Spaces and Operators ─────────────────────────
\subsubsection{Hilbert Spaces and Operators}

The mathematical framework of quantum mechanics is built on
\emph{Hilbert spaces}---complete inner product spaces that provide
the geometric structure needed to describe quantum states and their
transformations.

\begin{definition}[Inner product space]
\label{def:inner-product-space}
An \emph{inner product space} is a complex vector space~$H$ equipped
with a function
$\langle \cdot , \cdot \rangle : H \times H \to \mathbb{C}$
satisfying, for all $\psi, \phi, \chi \in H$ and
$\alpha, \beta \in \mathbb{C}$:
\begin{enumerate}
  \item \textbf{Conjugate symmetry.}
    $\langle \psi , \phi \rangle
    = \overline{\langle \phi , \psi \rangle}$.
  \item \textbf{Linearity in the second argument.}\footnote{This
    is the \emph{physics convention}.  The mathematics literature
    typically takes the inner product to be linear in the
    \emph{first} argument and conjugate-linear in the second.
    We follow the physics convention because it is standard in the
    quantum information literature and aligns with Dirac notation
    (introduced below).}
    $\langle \psi , \alpha\phi + \beta\chi \rangle
    = \alpha\langle \psi , \phi \rangle
      + \beta\langle \psi , \chi \rangle$.
  \item \textbf{Positive-definiteness.}
    $\langle \psi , \psi \rangle \geq 0$, with equality if and
    only if $\psi = 0$.
\end{enumerate}
The inner product induces a norm
$\|\psi\| = \sqrt{\langle \psi , \psi \rangle}$ and hence a
metric $d(\psi, \phi) = \|\psi - \phi\|$.
\end{definition}

\begin{definition}[Hilbert space]
\label{def:hilbert-space}
A \emph{Hilbert space} is an inner product space that is
\emph{complete}: every Cauchy sequence converges with respect to the
norm-induced metric.
\end{definition}

\noindent
Every finite-dimensional inner product space is automatically
complete,%
\footnote{A finite-dimensional complex inner product space is
isometrically isomorphic to~$\mathbb{C}^n$ (via any orthonormal
basis), and $\mathbb{C}^n$ is complete because $\mathbb{C}$ is
complete and a finite product of complete metric spaces is complete.}
so $\mathbb{C}^n$ with the standard inner product
$\langle \psi , \phi \rangle = \sum_i \overline{\psi_i}\,\phi_i$
is a Hilbert space for every~$n$.
Infinite-dimensional examples---such as the space~$\ell^2$ of
square-summable sequences---arise in the theory of continuous-variable
quantum systems, but all examples in this thesis are
finite-dimensional.

\paragraph{Dirac notation.}
Physicists conventionally write elements of a Hilbert space~$H$ as
\emph{kets} $\ket{\psi}$, emphasising their role as column vectors.
The corresponding \emph{bra} $\bra{\psi}$ is the conjugate-transpose
row vector, an element of the dual space~$H^*$.
The inner product of $\ket{\psi}$ and $\ket{\phi}$ is written
$\braket{\psi|\phi} \in \mathbb{C}$---a ``bracket'' split into a
bra and a ket.
The \emph{outer product} $\ket{\psi}\!\bra{\phi}$ is the linear
operator sending $\ket{\chi}$ to
$\braket{\phi|\chi}\,\ket{\psi}$.
We adopt Dirac notation throughout this subsection, as it is standard
in the quantum information
literature~\cite{NielsenChuang2010} and makes the interplay between
states and operators notationally transparent.

\begin{example}[The qubit Hilbert space]
\label{ex:qubit-hilbert}
The simplest quantum system is a single \emph{qubit}, modelled by the
Hilbert space $\mathbb{C}^2$.
The \emph{computational basis} consists of two orthonormal vectors
\[
  \ket{0} = \begin{pmatrix} 1 \\ 0 \end{pmatrix}, \qquad
  \ket{1} = \begin{pmatrix} 0 \\ 1 \end{pmatrix}.
\]
An arbitrary qubit state is a unit vector
$\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$ with
$|\alpha|^2 + |\beta|^2 = 1$.
The coefficients $\alpha, \beta \in \mathbb{C}$ are called
\emph{amplitudes}; their squared magnitudes give the probabilities
of observing $0$ or $1$ upon measurement.
\end{example}

\paragraph{Bounded operators.}
A linear map $A : H \to H$ is \emph{bounded} if there exists a
constant $c \geq 0$ such that
$\|A\ket{\psi}\| \leq c\,\|\ket{\psi}\|$ for
all $\ket{\psi} \in H$.
On finite-dimensional Hilbert spaces every linear map is automatically
bounded, but the distinction matters in the infinite-dimensional
setting.
We write $B(H)$ for the set of all bounded linear operators on~$H$.
Under operator addition, scalar multiplication, and composition,
$B(H)$ is an \emph{algebra}---a vector space equipped with an
associative bilinear product (here, operator
composition). This is the key mathematical structure on which
quantum order theory is built.

\begin{definition}[Adjoint]
\label{def:adjoint}
The \emph{adjoint} (or \emph{Hermitian conjugate}) of an operator
$A \in B(H)$ is the unique operator $A^\dagger \in B(H)$ satisfying
\[
  \langle \psi,\, A\phi \rangle
  = \langle A^\dagger\psi,\, \phi \rangle
\]
for all $\psi, \phi \in H$.
For finite-dimensional~$H$ with a chosen basis, the adjoint is the
conjugate transpose:
$(A^\dagger)_{ij} = \overline{A_{ji}}$.
An operator is \emph{self-adjoint} (or \emph{Hermitian}) if
$A = A^\dagger$.
\end{definition}

\begin{definition}[Positive semidefiniteness]
\label{def:positive-semidefinite}
An operator $A \in B(H)$ is \emph{positive semidefinite}, written
$A \geq 0$, if it is self-adjoint and
$\bra{\psi} A \ket{\psi} \geq 0$ for all $\ket{\psi} \in H$.
\end{definition}

\begin{example}[$2 \times 2$ operators]
\label{ex:2x2-operators}
On $\mathbb{C}^2$, the matrix
$A = \begin{psmallmatrix} 2 & 1 \\ 1 & 3 \end{psmallmatrix}$
is positive semidefinite: it is self-adjoint (its entries are real and
the matrix is symmetric) and its eigenvalues
$\frac{5 \pm \sqrt{5}}{2}$ are both positive.
The matrix
$B = \begin{psmallmatrix} 1 & 2 \\ 0 & 1 \end{psmallmatrix}$
is not self-adjoint ($B \neq B^\dagger$) and hence not positive
semidefinite.
The outer product
$\ket{0}\!\bra{0} =
\begin{psmallmatrix} 1 & 0 \\ 0 & 0 \end{psmallmatrix}$
is a \emph{projection}---a positive semidefinite operator satisfying
$P^2 = P$.
\end{example}

\begin{definition}[Trace]
\label{def:trace}
The \emph{trace} of an operator $A \in B(H)$
(for finite-dimensional~$H$) is the sum of its diagonal entries in
any orthonormal basis $\{\ket{e_i}\}$:
\[
  \tr(A) = \sum_i \bra{e_i} A \ket{e_i}.
\]
The trace is basis-independent, linear
($\tr(\alpha A + \beta B) =
\alpha\,\tr(A) + \beta\,\tr(B)$), and cyclic
($\tr(AB) = \tr(BA)$).
\end{definition}

\begin{definition}[L\"owner order]
\label{def:lowner-order}
The \emph{L\"owner order} on self-adjoint operators in $B(H)$ is
defined by
\[
  A \sqsubseteq B
  \quad\text{if and only if}\quad
  B - A \geq 0
  \quad\text{(i.e.\ $B - A$ is positive semidefinite)}.
\]
\end{definition}

\begin{example}[L\"owner order on $\mathbb{C}^2$]
\label{ex:lowner-2x2}
The identity matrix
$I = \begin{psmallmatrix} 1 & 0 \\ 0 & 1 \end{psmallmatrix}$
and the projection
$P = \ket{0}\!\bra{0} =
\begin{psmallmatrix} 1 & 0 \\ 0 & 0 \end{psmallmatrix}$
satisfy $P \sqsubseteq I$ in the L\"owner order, since
$I - P = \begin{psmallmatrix} 0 & 0 \\ 0 & 1 \end{psmallmatrix}
= \ket{1}\!\bra{1} \geq 0$.
More generally, every projection satisfies
$0 \sqsubseteq P \sqsubseteq I$.
\end{example}

\noindent
The L\"owner order makes the set of self-adjoint operators in $B(H)$
a partially ordered set.
It will reappear as the underlying order of the involutive quantale
$B(H)$ in \cref{def:inv-quantale}.

% ── Part 2: Quantum States and Computation ───────────────────────
\subsubsection{Quantum States and Computation}

With the linear-algebraic framework in place, we describe quantum
states and the basic operations of quantum computation.

\paragraph{Superposition.}
A quantum system associated with a Hilbert space~$H$ can exist in a
\emph{superposition} of basis states.
For a single qubit (\cref{ex:qubit-hilbert}), the state
$\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$ is not simply ``$0$
with probability~$|\alpha|^2$ and $1$ with
probability~$|\beta|^2$''.
It is a genuinely different kind of state that can exhibit
\emph{interference}---constructive or destructive combination of
amplitudes---with no classical analogue.
The superposition principle extends to systems of any dimension: any
unit vector in~$H$ is a valid quantum state.

\paragraph{Tensor products and multi-qubit systems.}
The state space of a composite quantum system is the \emph{tensor
product} of the component Hilbert spaces.
Two qubits together live in
$\mathbb{C}^2 \otimes \mathbb{C}^2 \cong \mathbb{C}^4$, with
computational basis
$\{\ket{00}, \ket{01}, \ket{10}, \ket{11}\}$,
where $\ket{ij} = \ket{i} \otimes \ket{j}$.
More generally, $n$~qubits live in
$(\mathbb{C}^2)^{\otimes n} \cong \mathbb{C}^{2^n}$---the
exponential growth of the state space is the source of quantum
computing's potential computational advantage.

\paragraph{Entanglement.}
A state of a composite system is \emph{separable} if it can be written
as a tensor product of states of the components, and \emph{entangled}
otherwise.

\begin{example}[Bell state]
\label{ex:bell-state}
The state
\[
  \ket{\Phi^+}
  = \frac{1}{\sqrt{2}}\bigl(\ket{00} + \ket{11}\bigr)
  \;\in\; \mathbb{C}^2 \otimes \mathbb{C}^2
\]
is entangled: there are no single-qubit states $\ket{a}, \ket{b}$
with $\ket{\Phi^+} = \ket{a} \otimes \ket{b}$.
Measuring either qubit in the computational basis yields $0$ or $1$
each with probability~$\tfrac{1}{2}$, but the outcomes are perfectly
correlated: if one qubit is measured as~$0$, the other is
guaranteed to be~$0$ as well, regardless of the spatial separation
between the two systems.
\end{example}

\noindent
Entanglement is a fundamentally quantum phenomenon with no classical
counterpart.
It plays a central role in quantum computation and, as we shall see,
is the reason that the positivity requirement for quantum channels
must be strengthened to \emph{complete} positivity
(\cref{def:cp-map}).

\paragraph{Measurement and the Born rule.}
The \emph{Born rule} governs the probabilistic outcomes of quantum
measurements.
For a system in state~$\ket{\psi}$ measured in an orthonormal basis
$\{\ket{e_i}\}$, the probability of outcome~$i$ is
$p_i = |\braket{e_i|\psi}|^2$,
and after obtaining outcome~$i$ the system collapses to the
post-measurement state~$\ket{e_i}$.
Measurement is irreversible: the superposition is destroyed and
cannot be recovered.

\begin{definition}[Density operator]
\label{def:density-operator}
A \emph{density operator} (or \emph{density matrix}) on a Hilbert
space~$H$ is an operator $\rho \in B(H)$ satisfying:
\begin{enumerate}
  \item \textbf{Positive semidefiniteness.}
    $\rho \geq 0$ (\cref{def:positive-semidefinite}).
  \item \textbf{Unit trace.} $\tr(\rho) = 1$ (\cref{def:trace}).
\end{enumerate}
\end{definition}

\noindent
A density operator is \emph{pure} if $\rho = \ket{\psi}\!\bra{\psi}$
for some unit vector $\ket{\psi} \in H$.
A \emph{mixed} density operator---one that is not pure---represents
statistical uncertainty about the quantum state.

\begin{example}[Qubit density operators]
\label{ex:qubit-density}
The pure state $\ket{0}$ corresponds to the density operator
$\rho_0 = \ket{0}\!\bra{0} =
\begin{psmallmatrix} 1 & 0 \\ 0 & 0 \end{psmallmatrix}$.
A classical coin flip between $\ket{0}$ and $\ket{1}$ is represented
by the mixed state
\[
  \rho_{\mathrm{mix}}
  = \tfrac{1}{2}\ket{0}\!\bra{0} + \tfrac{1}{2}\ket{1}\!\bra{1}
  = \begin{pmatrix} 1/2 & 0 \\ 0 & 1/2 \end{pmatrix}
  = \tfrac{1}{2}\,I.
\]
Both are positive semidefinite with unit trace, confirming that they
are valid density operators.
The mixed state $\rho_{\mathrm{mix}}$ cannot be written as
$\ket{\psi}\!\bra{\psi}$ for any~$\ket{\psi}$: it represents
genuine classical uncertainty.
\end{example}

\begin{remark}[Why density operators]
\label{rem:why-density}
Pure state vectors are insufficient for denotational semantics because
they cannot represent probabilistic mixtures or the partial
information obtained by observing a subsystem of an entangled system.
Density operators handle both: the Born rule generalises to
$p_i = \tr(P_i\,\rho)$ for measurement projectors~$P_i$, and the
\emph{partial trace} extracts subsystem states from entangled
composites.
For these reasons, the quantum information
literature---and this thesis---works primarily with density operators
rather than state vectors.
\end{remark}

% ── Part 3: Quantum Channels ─────────────────────────────────────
\subsubsection{Quantum Channels}

In classical domain theory, programs denote Scott-continuous functions
between CPOs (\cref{sec:bg-fixedpoints}).
In quantum computation, the analogous role is played by
\emph{quantum channels}: maps between spaces of operators that
respect the physical constraints of quantum mechanics.

\begin{definition}[Positive and completely positive maps]
\label{def:cp-map}
A linear map $\Phi : B(H) \to B(K)$ is:
\begin{enumerate}
  \item \textbf{Positive} if $\Phi(A) \geq 0$ whenever $A \geq 0$.
  \item \textbf{Completely positive (CP)} if, for every ancillary
    Hilbert space~$R$, the extended map
    $\Phi \otimes \mathrm{id}_R : B(H \otimes R) \to B(K \otimes R)$
    is positive.
\end{enumerate}
\end{definition}

\noindent
Positivity alone does not suffice for physically meaningful maps.
The issue is entanglement: a map~$\Phi$ may be applied to one half of
an entangled system, so the extended map
$\Phi \otimes \mathrm{id}_R$ must also send density operators to
density operators.

\begin{example}[A positive but not completely positive map]
\label{ex:transpose-not-cp}
The transpose map $T : A \mapsto A^T$ on $B(\mathbb{C}^2)$ is
positive: if $A \geq 0$ then $A^T \geq 0$ (since transposition
preserves eigenvalues).
However, applying $T \otimes \mathrm{id}$ to the density operator
$\ket{\Phi^+}\!\bra{\Phi^+}$ of the Bell state
(\cref{ex:bell-state}) yields an operator with a negative
eigenvalue---an unphysical result.
Complete positivity rules out such pathologies.
\end{example}

\begin{definition}[Trace-preserving map]
\label{def:tp-map}
A linear map $\Phi : B(H) \to B(K)$ is \emph{trace-preserving} (TP)
if $\tr(\Phi(A)) = \tr(A)$ for all $A \in B(H)$.
\end{definition}

\noindent
Trace preservation ensures that probabilities sum to one: if
$\rho$ is a density operator (trace one), then $\Phi(\rho)$ is also
trace one.

\begin{definition}[Quantum channel (CPTP map)]
\label{def:cptp-map}
A \emph{quantum channel} is a linear map
$\Phi : B(H) \to B(K)$ that is both completely positive and
trace-preserving (CPTP).
\end{definition}

\paragraph{Kraus representation.}
Every quantum channel admits a concrete representation that makes its
structure transparent.

\begin{theorem}[Kraus representation
  {\cite[Theorem~8.1]{NielsenChuang2010}}]
\label{thm:kraus}
A linear map $\Phi : B(H) \to B(K)$ is completely positive if and
only if there exist operators $\{K_i\}_{i=1}^r$ with
$K_i : H \to K$ such that
\[
  \Phi(\rho) = \sum_{i=1}^{r} K_i\,\rho\,K_i^\dagger.
\]
The map is additionally trace-preserving if and only if
$\sum_{i=1}^{r} K_i^\dagger K_i = I_H$.
The operators $\{K_i\}$ are called \emph{Kraus operators}.
\end{theorem}

\noindent
The Kraus representation makes the structure of a quantum channel
concrete: each Kraus operator describes one ``branch'' of the quantum
evolution, and the channel is the incoherent sum over all branches.
The trace-preservation constraint
$\sum_i K_i^\dagger K_i = I$ ensures that the total probability
across branches is one.

\begin{remark}[CPTP maps as morphisms]
\label{rem:cptp-morphisms}
The analogy with classical domain theory is direct:
\begin{center}
\begin{tabular}{@{}l l@{}}
\toprule
\textbf{Classical} & \textbf{Quantum} \\
\midrule
Domain (CPO) & Space of density operators on~$H$ \\
Scott-continuous function & CPTP map (quantum channel) \\
Preserves $\omega$-chain suprema & Preserves positivity and trace \\
\bottomrule
\end{tabular}
\end{center}
\noindent
In both settings, the morphisms preserve the structure that makes the
objects well-behaved semantic domains.
This parallel motivates the search for a unified categorical framework,
which $\CPO$-enrichment provides
(\cref{sec:disc-quantum-enrichment}).
\end{remark}

% ── Part 4: Quantum Control Flow ─────────────────────────────────
\subsubsection{Quantum Control Flow}

The preceding sections describe a world in which quantum data is
manipulated by channels, but all \emph{control decisions}---which
channel to apply, which branch to take---are classical.
This is the regime of most existing quantum programming languages,
including Selinger's QPL~\cite{Selinger2004},
the quantum lambda calculus of Selinger and
Valiron~\cite{SelingerValiron2009}, and their descendants.

\paragraph{Classical control over quantum data.}
In a language with \emph{classical control flow}, a branching
construct
\[
  \mathtt{if}\;b\;\mathtt{then}\;\Phi_0\;\mathtt{else}\;\Phi_1
\]
requires the guard~$b$ to be a classical bit---typically the result of
a prior measurement.
The choice of which channel to apply is determined \emph{before} the
quantum data is touched.
Classical control suffices for standard quantum algorithms (Shor's
algorithm, Grover's search, quantum error correction), but it does not
exhaust the possibilities of quantum mechanics.
\Cref{fig:classical-control} illustrates the structure.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    >=Stealth,
    box/.style={draw, minimum width=1.4cm, minimum height=0.8cm,
                font=\small},
    diamond/.style={draw, shape=diamond, aspect=2, minimum width=1.2cm,
                    inner sep=1pt, font=\small},
    every edge/.style={draw, ->},
  ]
  % Input
  \node[box] (rho) {$\rho$};
  % Decision
  \node[diamond, right=1.1cm of rho] (b) {$b$};
  % Branches
  \node[box, above right=0.6cm and 1.0cm of b] (phi0) {$\Phi_0$};
  \node[box, below right=0.6cm and 1.0cm of b] (phi1) {$\Phi_1$};
  % Merge
  \node[coordinate, right=1.0cm of phi0] (m0) {};
  \node[coordinate, right=1.0cm of phi1] (m1) {};
  \node[coordinate] (merge) at ($(m0)!0.5!(m1)$) {};
  % Output
  \node[box, right=0.4cm of merge] (out) {$\rho'$};
  % Edges
  \draw[->] (rho) -- (b);
  \draw[->] (b) -- node[above left, font=\scriptsize] {$0$} (phi0);
  \draw[->] (b) -- node[below left, font=\scriptsize] {$1$} (phi1);
  \draw[->] (phi0) -- (m0) -- (merge);
  \draw[->] (phi1) -- (m1) -- (merge);
  \draw[->] (merge) -- (out);
\end{tikzpicture}
\caption{Classical control flow: the classical bit~$b$ (obtained by
  measurement) selects exactly one of $\Phi_0$ or~$\Phi_1$.
  The two branches are mutually exclusive.}
\label{fig:classical-control}
\end{figure}

\paragraph{Quantum control flow.}
\emph{Quantum control flow} allows the guard to be an unmeasured
qubit in superposition: the ``choice'' of which branch to take is
itself a quantum degree of freedom.
Both branches coexist coherently, and the system can exhibit
interference between them.
The canonical example is the \emph{quantum switch}~\cite{CDPV2013}.

\paragraph{The quantum switch.}
Given two quantum channels $A$ and $B$ acting on a target system with
Hilbert space~$H$, and a control qubit in state
$\ket{c} = \alpha\ket{0} + \beta\ket{1}$, the quantum switch
produces a channel on the composite system
$H \otimes \mathbb{C}^2$ that applies $A$ then~$B$ to the target on
the $\ket{0}$ component of the control and $B$ then~$A$ on the
$\ket{1}$ component---\emph{coherently}, without measuring the
control qubit.
When the control is in superposition, the result is a genuine
superposition of causal orders: the target system experiences both
$A$-then-$B$ and $B$-then-$A$ simultaneously, with quantum
interference between the two orderings~\cite{CDPV2013}.
This process has no classical analogue---there is no classical
computation that produces a ``superposition of the order in which two
operations are applied''.
The quantum switch has been demonstrated
experimentally~\cite{RichterEtAl2025}.
\Cref{fig:quantum-switch} illustrates the structure; contrast with
the classical branching of \cref{fig:classical-control}.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    >=Stealth,
    box/.style={draw, minimum width=1.4cm, minimum height=0.8cm,
                font=\small},
    every edge/.style={draw, ->},
  ]
  % Switch box (outer)
  \node[draw, dashed, rounded corners=6pt,
        minimum width=4.2cm, minimum height=3.0cm,
        label={[font=\small]above:quantum switch}] (sw) {};
  % Inner paths — stacked
  \node[box, fill=blue!8] (a1) at ($(sw.center)+(-.7cm, 0.55cm)$) {$A$};
  \node[box, fill=blue!8] (b1) at ($(sw.center)+( .7cm, 0.55cm)$) {$B$};
  \node[box, fill=red!8]  (b2) at ($(sw.center)+(-.7cm,-0.55cm)$) {$B$};
  \node[box, fill=red!8]  (a2) at ($(sw.center)+( .7cm,-0.55cm)$) {$A$};
  % Path labels
  \node[font=\scriptsize, left=0.05cm of a1, blue!70!black] {$\ket{0}$:};
  \node[font=\scriptsize, left=0.05cm of b2, red!70!black]  {$\ket{1}$:};
  % Arrows inside
  \draw[->, blue!70!black]  (a1) -- (b1);
  \draw[->, red!70!black]   (b2) -- (a2);
  % Input
  \node[box, left=0.8cm of sw] (rho2) {$\rho$};
  \node[font=\small, below=0.2cm of rho2] {$\ket{c} = \alpha\ket{0}+\beta\ket{1}$};
  % Output
  \node[box, right=0.8cm of sw] (out2) {$\rho'$};
  % Edges
  \draw[->] (rho2) -- (sw);
  \draw[->] (sw)   -- (out2);
\end{tikzpicture}
\caption{The quantum switch: both causal orders coexist coherently,
  controlled by the unmeasured qubit~$\ket{c}$.
  The dashed boundary indicates a single coherent process, not a
  sequential circuit---no measurement or branching decision occurs.}
\label{fig:quantum-switch}
\end{figure}

\paragraph{The insufficiency of Boolean order.}
The quantum switch exposes a fundamental limitation of classical
domain theory.
In a classical CPO, the order relation $x \sqsubseteq y$ is a
proposition in~$\mathsf{Prop}$: it is simply true or false.
A semantic model built on such an order can describe a world in which
one branch \emph{or} the other is taken, but not a coherent
superposition of both.
To model quantum control flow, we need an order relation that can
carry \emph{quantum information}---not a truth value in
$\{0, 1\}$, but an operator in~$B(H)$.
This is the conceptual motivation for quantale-valued order.

% ── Part 5: From Boolean to Quantale-Valued Order ────────────────
\subsubsection{From Boolean to Quantale-Valued Order}

The preceding subsection identified the obstacle: classical order
relations are Boolean-valued and cannot carry quantum information.
The solution, due to Kornell, Lindenhovius, and
Mislove~\cite{KLM2024} (building on Weaver~\cite{Weaver2010}),
replaces the Boolean truth value with an element of an
\emph{involutive quantale}.

\paragraph{Quantales informally.}
A quantale is, informally, a complete lattice equipped with a
compatible monoid structure.
The lattice captures ``how much truth'' a statement carries---generalising
from the two-element Boolean lattice $\{0, 1\}$ to a richer
ordered structure---while the monoid operation combines evidence,
analogous to conjunction.
An \emph{involutive} quantale additionally has an involution
$(-)^*$, mirroring the adjoint operation on~$B(H)$.

\begin{definition}[Involutive quantale]
\label{def:inv-quantale}
An \emph{involutive quantale} is a tuple
$(Q, \leq, \top, \bot, \otimes, (-)^*, \wedge, \bigwedge)$ where:
\begin{enumerate}
  \item $(Q, \leq)$ is a partial order with top~$\top$,
    bottom~$\bot$, binary meet~$\wedge$, and arbitrary
    infima~$\bigwedge$.
  \item $\otimes : Q \times Q \to Q$ is an associative binary product
    with unit~$\top$ that is monotone in each argument.
  \item $(-)^* : Q \to Q$ is an order-reversing involution
    satisfying $(a \otimes b)^* = b^* \otimes a^*$.
\end{enumerate}
\end{definition}

\begin{example}[The Boolean quantale]
\label{ex:boolean-quantale}
The two-element set $\{\bot, \top\}$ with the usual order,
$\otimes = \wedge$ (conjunction), and the identity involution is an
involutive quantale.
This is the quantale that underlies classical order theory: an order
relation $x \sqsubseteq y$ valued in $\{\bot, \top\}$ is simply a
proposition that is true or false.
\end{example}

\begin{example}[$B(H)$ as an involutive quantale]
\label{ex:bh-quantale}
For a Hilbert space~$H$, the algebra $B(H)$ of bounded operators
forms an involutive quantale with:
\begin{itemize}
  \item $\leq$ the L\"owner order (\cref{def:lowner-order});
  \item $\top = I$ (the identity operator) and $\bot = 0$ (the zero
    operator);
  \item $\otimes$ operator composition;
  \item $(-)^*$ the adjoint (\cref{def:adjoint});
  \item $\wedge$ and $\bigwedge$ the infima in the L\"owner order.
\end{itemize}
When $H = \mathbb{C}^2$, this is the algebra of
$2 \times 2$ complex matrices.
An order relation valued in $B(\mathbb{C}^2)$ assigns to each pair
$(x, y)$ a $2 \times 2$ matrix $R(x, y)$---not a truth value, but a
quantum operator encoding how much information witnesses the
approximation $x \sqsubseteq y$.
\end{example}

\noindent
With the quantale structure in hand, we define the quantum
generalisation of a partial order.

\begin{definition}[Quantum poset {\cite[Definition~2.6.1]{KLM2024}}]
\label{def:qposet}
Let $Q$ be an involutive quantale.
A \emph{$Q$-valued quantum poset} is a set~$X$ with decidable
equality, equipped with a function $R : X \times X \to Q$ satisfying:
\begin{enumerate}
  \item \textbf{Reflexivity.} $\top \leq R(x, x)$ for all~$x$.
  \item \textbf{Transitivity.}
    $R(x, y) \otimes R(y, z) \leq R(x, z)$ for all~$x, y, z$.
  \item \textbf{Antisymmetry.}
    $R(x, y) \wedge R(y, x) \leq \delta(x, y)$, where
    $\delta$ is the Kronecker delta
    ($\delta(x,x) = \top$, $\delta(x,y) = \bot$ for $x \neq y$).
\end{enumerate}
\end{definition}

\noindent
When $Q$ is the Boolean quantale (\cref{ex:boolean-quantale}), a
quantum poset reduces to an ordinary partial order: the function~$R$
becomes a Boolean predicate, the quantale product
$\otimes = \wedge$ becomes conjunction, and the three axioms
specialise to the standard reflexivity, transitivity, and
antisymmetry.

The quantale-valued formulation generalises the order relation from a
truth value to a \emph{quantum measurement outcome}: $R(x, y)$
encodes not just \emph{whether} $x$ approximates~$y$, but
\emph{how much quantum information} witnessing the approximation is
available.
This is the foundation on which quantum CPOs and the quantum Kleene
fixed-point theorem are built (\cref{sec:disc-qcpo}).

% ·· Formalisation ···········································
\subsection{Quantum CPOs and Quantum Morphisms}
\label{sec:disc-qcpo}

With the mathematical context established, we describe our
formalisation of the quantum domain theory of Kornell, Lindenhovius,
and Mislove~\cite{KLM2024}.
The implementation comprises three files totalling 1{,}749 lines of
substantive \Rocq{} code (\texttt{QuantumStructure.v},
\texttt{qCPO.v}, \texttt{qCPOProperties.v}), plus two stub files
reserved for future extensions.

\paragraph{Involutive quantales.}
The quantale~$Q$ is formalised as an \HB{} structure.
The mixin \texttt{HasQuantaleOps} provides the six operations
($\top$, $\bot$, $\otimes$, $(-)^*$, $\wedge$, $\bigwedge$),
and \texttt{IsInvQuantale} asserts the fourteen axioms of
\cref{def:inv-quantale} grouped into five families: extremal
elements, product laws, involution laws, meet laws, and infimum laws.
Using \HB{} here has a concrete payoff: any type registered as an
\texttt{InvQuantale.type} automatically inherits partial-order
structure, and the elaborator resolves notational overloading
(e.g.\ $\leq$ for the quantale order) without explicit annotations.

\paragraph{Quantum posets.}
Quantum posets (\cref{def:qposet}) are formalised as a plain
\texttt{Record qposet Q} parameterised by an involutive quantale~$Q$.
The record bundles a carrier type~$X$, the $Q$-valued order
$R : X \to X \to Q$, decidable equality on~$X$, and the three axioms
(reflexivity, transitivity, antisymmetry).
Plain records are used rather than \HB{} structures because \HB{}
does not handle structures parameterised by other structures well
(DD-022 in \cref{app:design-decisions}); this follows the same
pattern as \texttt{chain}, \texttt{mono\_fun}, and \texttt{cont\_fun}
in the classical layer.

\begin{lstlisting}[caption={The \texttt{qposet} record (\texttt{QuantumStructure.v}).},
                    label={lst:qposet}]
Record qposet (Q : InvQuantale.type) : Type := Build_qposet {
    qp_carrier :> Type;
    qp_ord     : qp_carrier -> qp_carrier -> Q;
    qp_dec_eq  : forall (x y : qp_carrier), {x = y} + {x <> y};
    (* Reflexivity: q_top ⊑ R(x, x) *)
    qp_refl    : forall (x : qp_carrier),
        q_top ⊑ qp_ord x x;
    (* Transitivity: R(x,y) ⊗ R(y,z) ⊑ R(x,z) *)
    qp_trans   : forall (x y z : qp_carrier),
        qp_ord x y ⊗ qp_ord y z ⊑ qp_ord x z;
    (* Antisymmetry: R(x,y) ⊓ q_adj(R(y,x)) ⊑ q_delta(x,y) *)
    qp_antisym : forall (x y : qp_carrier),
        qp_ord x y ⊓ q_adj (qp_ord y x) ⊑ q_delta qp_dec_eq x y;
}.
\end{lstlisting}

\noindent
The three axioms are direct translations of
\cref{def:qposet}: reflexivity replaces
$\top \leq R(x, x)$ with \lstinline{q_top ⊑ qp_ord x x},
transitivity replaces conjunction with the quantale tensor product
$\otimes$, and antisymmetry uses the quantale meet~$\sqcap$, the
involution~$(-)^*$, and the Kronecker delta \texttt{q\_delta}
(which returns $\top$ when $x = y$ and $\bot$ otherwise,
using the decidable equality \texttt{qp\_dec\_eq}).

\paragraph{Quantum chains and convergence.}
A \emph{quantum chain} in a quantum poset~$X$ is a family
$K : \mathbb{N} \to W \to X$ (parameterised by an arbitrary probe
type~$W$) that is ascending:
$\top \leq R(K(n, w),\; K(n{+}1, w))$ for all~$n$ and~$w$.
A function $K_\infty : W \to X$ is the \emph{limit} of~$K$ when
\[
  R(K_\infty(w),\; x) \;=\; \textstyle\bigwedge_n\, R(K(n, w),\; x)
\]
for all $w$ and~$x$---that is, the $Q$-valued distance from the limit
to any point equals the infimum of the distances from the chain
elements~\cite[Definition~3.1.1]{KLM2024}.
In \Rocq{}, this equality is split into two directions:

\begin{lstlisting}[caption={Quantum-chain convergence (\texttt{qCPO.v}).},
                    label={lst:converges}]
Definition converges (K : qchain) (K_inf : W -> X) : Prop :=
    forall (w : W) (x : X),
        qp_ord X (K_inf w) x ⊑ q_inf (qord_desc_chain K w x)
        /\
        q_inf (qord_desc_chain K w x) ⊑ qp_ord X (K_inf w) x.
\end{lstlisting}

\noindent
The auxiliary \texttt{qord\_desc\_chain K w x} packages the
sequence $n \mapsto R(K(n, w),\; x)$ as a descending $\omega$-chain
in~$Q$---descending because $R$ is antitone in its first argument
when the chain is ascending---so that \texttt{q\_inf} can compute
its greatest lower bound.

\begin{definition}[Quantum CPO {\cite[Definition~3.2.1]{KLM2024}}]
\label{def:qcpo}
A quantum poset~$X$ is a \emph{quantum $\omega$-CPO}
(with respect to probe type~$W$)
if every quantum $W$-chain in~$X$ has a limit.
A \emph{pointed quantum CPO} additionally has a bottom element
$\bot_X$ satisfying $\top \leq R(\bot_X, x)$ for all~$x$.
\end{definition}

\noindent
The formalisation provides both a propositional version
(\texttt{is\_qcpo}) and a data-carrying version (\texttt{QCPOData})
that bundles the limit function with a convergence proof, mirroring
the \texttt{HasSup}/\texttt{IsCPO} split in the classical hierarchy.

\paragraph{Quantum-continuous morphisms.}
A function $F : X \to Y$ between quantum posets is
\emph{quantum-monotone} if
$R_X(x, x') \leq R_Y(F(x), F(x'))$
and \emph{quantum Scott-continuous} if additionally
$F$ preserves limits of quantum chains.
These are bundled in a \texttt{q\_cont\_fun} record with three
fields: the function, a monotonicity proof, and a
convergence-preservation proof.
The record supports identity, composition (with separate
monotonicity and preservation proofs composed independently), and
the three category laws (left/right identity, associativity)---all
proved using \texttt{FunctionalExtensionality} and
\texttt{ProofIrrelevance}, the same strategy as the classical
\texttt{cont\_fun} (\cref{sec:meth-function-spaces}).

\begin{lstlisting}[caption={Bundled quantum Scott-continuous maps
  (\texttt{qCPOProperties.v}).},
                    label={lst:qcontfun}]
Record q_cont_fun (X Y : qposet Q) : Type := Build_q_cont_fun {
    qcf_fun :> X -> Y;
    qcf_mono : q_monotone qcf_fun;
    qcf_preserves : forall (W : Type)
        (K : @qchain Q X W) (K_inf : W -> X),
        converges K K_inf ->
        converges (map_qchain qcf_fun qcf_mono K)
                  (fun w => qcf_fun (K_inf w));
}.
\end{lstlisting}

\noindent
The split-field design (DD-023 in \cref{app:design-decisions}) keeps
monotonicity and convergence-preservation as separate record fields
rather than embedding monotonicity inside an existential.  This
ensures that the monotonicity witness is always directly available
for \texttt{map\_qchain}---the function that applies a monotone
map pointwise to a quantum chain---avoiding proof-witness mismatches
during composition.

\paragraph{The quantum Kleene fixed-point theorem.}
Given a pointed quantum CPO~$X$ with bottom~$\bot_X$ and a
quantum Scott-continuous endomorphism $F : X \to X$, the sequence
$\bot_X, F(\bot_X), F^2(\bot_X), \ldots$ forms an ascending quantum
chain.
Its limit is the least fixed point of~$F$:
\begin{enumerate}
  \item $R(F(\mathrm{fix}), y) = R(\mathrm{fix}, y)$
    for all~$y$---i.e.\ $F(\mathrm{fix})$ is indistinguishable
    from~$\mathrm{fix}$ in the quantum order.
  \item For any prefixed point~$x$
    ($\top \leq R(F(x), x)$),
    $\top \leq R(\mathrm{fix}, x)$---i.e.\ $\mathrm{fix}$ is least.
\end{enumerate}
The proof (\texttt{qfixp\_at\_is\_fixedpoint},
\texttt{qfixp\_at\_least} in \texttt{qCPOProperties.v})
mirrors the classical Kleene construction
(\cref{sec:bg-fixedpoints}), with $Q$-valued inequalities
replacing Boolean ones throughout.
A quantum Scott induction principle (\texttt{qfixp\_at\_ind}) is
also derived: any admissible predicate that holds at~$\bot_X$ and is
preserved by~$F$ holds at $\mathrm{fix}$.

\begin{lstlisting}[caption={Kleene iteration and ascending chain
  (\texttt{qCPOProperties.v}).},
                    label={lst:qiter}]
Fixpoint q_iter (F : [X →qc X]) (n : nat) : X :=
    match n with
    | O   => q_bot_elem bot
    | S k => F (q_iter F k)
    end.

Lemma q_iter_ascending (F : [X →qc X]) (n : nat) (w : W) :
    (q_top : Q) ⊑ qp_ord X (q_iter F n) (q_iter F (S n)).
\end{lstlisting}

\noindent
The iteration has the same structure as its classical counterpart
(\cref{sec:bg-fixedpoints}): the base case is the bottom element,
and each step applies~$F$.  The ascending lemma shows that each
consecutive pair is related in the quantum order; the proof proceeds
by induction, using the bottom property at the base case and
monotonicity of~$F$ at the step.

% ·· Enrichment ··············································
\subsection{The Enrichment Correspondence}
\label{sec:disc-quantum-enrichment}

The central structural observation of this thesis is that the
category of quantum CPOs is $\CPO$-enriched.
This subsection explains what that means concretely and what the
formalisation establishes.

\paragraph{Hom-sets as CPOs.}
Given quantum CPOs $X$ and $Y$, the quantum Scott-continuous maps
$[X \to_q Y]$ carry a pointwise quantum order:
\[
  f \sqsubseteq g
  \;\;\text{iff}\;\;
  \forall\, x.\;
  \top \leq R_Y(f(x),\; g(x)).
\]

\begin{lstlisting}[caption={Pointwise ordering on quantum hom-sets
  (\texttt{qCPOProperties.v}).},
                    label={lst:qhomle}]
Definition q_hom_le (f g : [X →qc Y]) : Prop :=
    forall (x : X),
        (q_top : Q) ⊑ qp_ord Y (qcf_fun f x) (qcf_fun g x).
\end{lstlisting}

\noindent
This order is reflexive, transitive, and antisymmetric (the last by
extensionality of \texttt{q\_cont\_fun}), making
$[X \to_q Y]$ a partial order.
Ascendingness of a chain of morphisms $(f_n)$ is inherited
pointwise, and the limit is constructed as the pointwise limit via
\texttt{QCPOData}:
$(\bigsqcup_n f_n)(x) = \bigsqcup_n f_n(x)$.
The proof that this limit is itself quantum Scott-continuous
(\texttt{qhom\_limit\_cont}, 150 lines) is the technical core of
the quantum layer: it requires showing that the pointwise limit of
continuous maps preserves convergence of arbitrary quantum chains, a
three-step argument analogous to the classical
\texttt{fun\_sup}~(\cref{sec:meth-function-spaces}).

This means $[X \to_q Y]$ is an $\omega$-\textbf{CPO} (a classical
$\omega$-CPO, not merely a quantum one).
This is precisely the $\CPO$-enrichment condition
(\cref{def:cpo-enriched-cat}): the hom-objects are CPOs, and
composition is Scott-continuous in each variable---the latter
following from the separate continuity of quantum function
composition.

\paragraph{The unifying correspondence.}
Both the classical category $\CPO$ and the quantum category $\qCPO$
satisfy the \texttt{CPOEnrichedCat} axioms:
\begin{itemize}
  \item In $\CPO$: hom-sets $[D \to_c E]$ are CPOs
    (\cref{sec:meth-function-spaces}), and composition is separately
    Scott-continuous (DD-002 in \cref{app:design-decisions}).
  \item In $\qCPO$: hom-sets $[X \to_q Y]$ are CPOs (this section),
    and composition is separately continuous
    (\texttt{q\_cont\_comp} in \texttt{qCPOProperties.v}).
\end{itemize}
Every theorem proved generically over a \texttt{CPOEnrichedCat}---locally continuous functors, enriched natural transformations with
pointwise CPO structure, the Yoneda lemma, recursive domain equations
via bilimits---therefore applies to quantum domains without
modification, provided the enrichment instance is registered.

\paragraph{The gap.}
The formalisation demonstrates that all the required properties
hold (\texttt{q\_hom\_le\_refl}, \texttt{q\_hom\_le\_trans},
\texttt{q\_hom\_le\_antisym\_ord}, \texttt{qhom\_limit\_cont},
and the category laws in \texttt{qCPOProperties.v}), but does
\emph{not} yet register $\qCPO$ as an \texttt{HB.instance} of
\texttt{CPOEnrichedCat}.
The obstacle is engineering, not mathematics: \HB{} requires the
objects and morphisms to be packaged as \HB{}-compatible types,
whereas the quantum layer uses plain records parameterised by a
quantale (\cref{sec:disc-tradeoffs}).
Concretely, the \texttt{q\_cont\_fun} record would need to be
promoted to an \HB{} structure---or a wrapper type
introduced---before the instance declaration compiles.
The two stub files \texttt{QuantumMorphisms.v} and
\texttt{QuantumEnrichment.v} mark where this bridge would be
built.

\paragraph{Limitations and future directions.}
The quantum layer is exploratory.
Three significant gaps remain beyond the enrichment instance:
\begin{enumerate}
  \item \textbf{No concrete quantale instance.}
    The library axiomatises $Q$ as an abstract
    \texttt{InvQuantale.type} but does not construct the concrete
    example $Q = B(H)$.
    Doing so would require a Hilbert-space formalisation not available
    in the current \Rocq{} ecosystem.

  \item \textbf{No quantum lift monad.}
    The classical lift $D_\bot = \texttt{option}\;D$ has no direct
    quantum analogue; the quantum variant would need to interact with
    the quantale-valued order in a way that is not yet clear.

  \item \textbf{No quantum language.}
    The stub files \texttt{QMiniCore\_Syntax.v} and
    \texttt{QMiniCore\_Semantics.v} in \texttt{lang/} are reserved
    for a quantum programming language whose types would be
    interpreted as quantum CPOs, but no syntax or semantics has been
    specified.
\end{enumerate}
These gaps are discussed further in the future-work section of
\cref{ch:conclusion}.

% ── The lift monad obstruction ────────────────────────────────────
\section{The Coinductive Lift Monad}
\label{sec:disc-lift-monad}

% Source: theory/LiftMonad.v

The library uses the flat lift $D_\bot = \texttt{option}\;D$ to model
partiality (\cref{sec:meth-lift}).
The original BKV library~\cite{BKV2009} used instead a coinductive
\texttt{delay} type, which more faithfully models the step-by-step
unfolding of a partial computation.
This section explains why the coinductive lift cannot be made into a
\texttt{CPO.type} in our Leibniz-equality framework, develops the
obstruction formally in \Rocq{}, and discusses the resolution.
\Cref{sec:disc-lift-bg} introduces coinduction and the delay monad,
\cref{sec:disc-lift-obstruction} states the CPO obstruction theorem,
and \cref{sec:disc-lift-resolution} discusses the design trade-off.

% ·· Background ··············································
\subsection{Background: Coinduction and the Delay Monad}
\label{sec:disc-lift-bg}

This subsection introduces \emph{coinductive types} and the
\emph{delay monad}---the concepts needed to understand the CPO
obstruction developed in \cref{sec:disc-lift-obstruction}.
The reader familiar with coinduction in \Rocq{} may skip to
\cref{sec:disc-lift-obstruction}.

\paragraph{Coinductive types.}
\Rocq{}'s \texttt{Inductive} types define the \emph{smallest} type closed
under a set of constructors: every element is a finite tree built from
those constructors, and elimination proceeds by structural recursion.
\emph{Coinductive} types, declared with \texttt{CoInductive}, dualise
this: they define the \emph{largest} type closed under the constructors,
admitting potentially infinite inhabitants.
Where an inductive type is characterised by how its elements are
\emph{built} (introduction), a coinductive type is characterised by
how its elements are \emph{observed} (elimination by pattern matching,
one constructor at a time).
The canonical example is a stream $\texttt{Stream}\;A$: an infinite
sequence of $A$-values, destructed into a head and a tail but never
fully traversed.

\paragraph{Guardedness.}
Functions producing coinductive values are defined with
\texttt{CoFixpoint} rather than \texttt{Fixpoint}.
Whereas recursive functions must be \emph{structurally decreasing} on
an inductive argument, corecursive functions must be
\emph{guarded}: every corecursive call must appear directly under a
constructor of the coinductive type.
This ensures that the output can always be observed one constructor
at a time---\Rocq{}'s productivity guarantee, the dual of
termination for inductive recursion.

The guardedness requirement interacts awkwardly with \Rocq{}'s
reduction strategy.
A \texttt{CoFixpoint} definition does not reduce at the top level;
it only unfolds when it is the scrutinee of a \texttt{match}.
This means that a propositional equality like
$\texttt{bind}\;(\texttt{now}\;d)\;f = f\;d$---which should hold by
computation---is \emph{not} provable by \texttt{reflexivity},
because the left-hand side is a stuck cofixpoint.
The standard workaround is the \emph{frob trick}%
~\cite{Capretta2005}: define a function
\texttt{delay\_decompose} that pattern-matches on a \texttt{delay}
value and reconstructs it, then use the propositional identity
$t = \texttt{delay\_decompose}\;t$ to force one step of cofixpoint
evaluation.
This is a routine but pervasive annoyance in coinductive developments.

\paragraph{The delay type.}
The \emph{delay monad}, introduced by
Capretta~\cite{Capretta2005}, models partial computations as
coinductive sequences of ``computation steps'':

\begin{lstlisting}[caption={The coinductive delay type
  (\texttt{LiftMonad.v}).},
                    label={lst:delay}]
CoInductive delay (D : Type) : Type :=
  | now   : D -> delay D
  | later : delay D -> delay D.
\end{lstlisting}

\noindent
A value $\texttt{now}\;d$ represents a terminated computation
returning~$d$.
A value $\texttt{later}\;t$ represents one more step of
computation before~$t$.
The bottom element is the infinite loop
$\bot = \texttt{cofix}\;\bot := \texttt{later}\;\bot$, which
unfolds to $\texttt{later}\;(\texttt{later}\;(\texttt{later}\;
\cdots))$ and never produces a value.

This type has a natural monad structure.
The unit is $\texttt{now} : D \to \texttt{delay}\;D$, and the bind
operation sequences computations:
\texttt{bind} peels off \texttt{later} constructors until it
reaches a \texttt{now}\;$d$, then applies the continuation.
The three monad laws (left unit, right unit, associativity) all hold
\emph{up to bisimulation}---not up to propositional equality, for
the reduction reasons described above.

\paragraph{Weak bisimulation.}
Propositional equality is the wrong notion of equivalence on
\texttt{delay}\;$D$.
The terms $\texttt{now}\;d$ and $\texttt{later}\;(\texttt{now}\;d)$
are propositionally distinct (they have different outermost
constructors), but they should be semantically equivalent: one extra
computation step makes no observable difference.
The correct equivalence is \emph{weak bisimulation}: a coinductive
relation that allows either side to take arbitrarily many
\texttt{later} steps before matching.

\begin{lstlisting}[caption={Weak bisimulation on \texttt{delay}
  (\texttt{LiftMonad.v}).},
                    label={lst:bisim}]
CoInductive bisim {D : Type} : delay D -> delay D -> Prop :=
  | bisim_now   : forall d,     bisim (now d) (now d)
  | bisim_later : forall t1 t2, bisim t1 t2 ->
                                bisim (later t1) (later t2)
  | bisim_left  : forall t1 t2, bisim t1 t2 ->
                                bisim (later t1) t2
  | bisim_right : forall t1 t2, bisim t1 t2 ->
                                bisim t1 (later t2).
\end{lstlisting}

\noindent
The four constructors capture the idea that \texttt{later} steps on
either side can be silently absorbed.
\texttt{bisim\_now} is the base case: two terminated computations
with the same value are bisimilar.
\texttt{bisim\_later} steps both sides simultaneously.
\texttt{bisim\_left} and \texttt{bisim\_right} skip a step on one
side only.

Bisimulation is reflexive (by corecursion on the \texttt{delay}
structure) and symmetric (by swapping \texttt{bisim\_left} and
\texttt{bisim\_right}).
Transitivity, however, is problematic---this point is taken up in
\cref{sec:disc-lift-obstruction}.

\paragraph{Convergence.}
The \emph{convergence} relation is the dual concept: an
\emph{inductive} (finitary) predicate asserting that a computation
terminates with a specific value after finitely many steps.

\begin{definition}[Convergence]
\label{def:convergence}
$\texttt{converges}\;t\;d$ holds if $t$ reaches
$\texttt{now}\;d$ after finitely many \texttt{later} steps:
\begin{enumerate}
  \item $\texttt{converges}\;(\texttt{now}\;d)\;d$, and
  \item if $\texttt{converges}\;t\;d$ then
    $\texttt{converges}\;(\texttt{later}\;t)\;d$.
\end{enumerate}
Convergence is functional: if $\texttt{converges}\;t\;d$ and
$\texttt{converges}\;t\;d'$, then $d = d'$.
The bottom element $\bot$ does not converge to any value.
\end{definition}

\noindent
The interaction between coinductive bisimulation and inductive
convergence is where the technical difficulties arise, as we now
describe.

% ·· The CPO obstruction ·····································
\subsection{The CPO Obstruction}
\label{sec:disc-lift-obstruction}

The BKV library~\cite{BKV2009} builds on a preorder hierarchy where
two elements can be mutually below without being propositionally
equal (\cref{sec:disc-bkv}).
Our library requires Leibniz antisymmetry:
$x \sqsubseteq y \land y \sqsubseteq x \implies x = y$.
This section shows that no ordering on \texttt{delay}\;$D$ can
simultaneously respect the convergence structure and satisfy Leibniz
antisymmetry.

\paragraph{The convergence preorder.}
The natural ordering on \texttt{delay}\;$D$ is the
\emph{convergence preorder}:
\[
  t \sqsubseteq t'
  \quad\text{iff}\quad
  \forall\, d.\;
  \texttt{converges}\;t\;d \implies \texttt{converges}\;t'\;d.
\]
A computation $t'$ is ``at least as defined as''~$t$ if every value
that $t$ produces is also produced by~$t'$.
This is a preorder (reflexive and transitive), but it is not
antisymmetric under propositional equality.
The obstruction is captured by three facts, which together constitute
the main result of \texttt{LiftMonad.v}:

\begin{lstlisting}[caption={The CPO obstruction theorem
  (\texttt{LiftMonad.v}).},
                    label={lst:obstruction}]
(*  FACT A: bisimilar (should be equal) *)
Fact A {D : Type} (d : D) :
    bisim (now d) (later (now d)).

(*  FACT B: propositionally unequal *)
Fact B {D : Type} (d : D) :
    now d <> later (now d).

(*  FACT C: mutually below in convergence order *)
Fact C {D : Type} (d : D) :
    forall e,
        converges (now d) e <->
        converges (later (now d)) e.

Theorem delay_CPO_requires_quotient {D : Type} (d : D) :
    bisim (now d) (later (now d)) /\
    now d <> later (now d) /\
    (forall e,
        converges (now d) e <->
        converges (later (now d)) e).
\end{lstlisting}

\noindent
\textbf{Fact~A} states that $\texttt{now}\;d$ and
$\texttt{later}\;(\texttt{now}\;d)$ are bisimilar: the extra
\texttt{later} step is absorbed by \texttt{bisim\_right}.
In any CPO modelling partial computation, these two elements
\emph{should} be identified---they represent the same
computation, one finishing immediately and the other taking one
extra step.

\textbf{Fact~B} states that $\texttt{now}\;d \neq
\texttt{later}\;(\texttt{now}\;d)$ in propositional equality.
The proof is a single tactic: \texttt{discriminate}, which exploits
the fact that distinct constructors of an inductive (or coinductive)
type yield distinct terms.

\textbf{Fact~C} states that these two terms are mutually below in
the convergence order: anything that $\texttt{now}\;d$ converges to,
$\texttt{later}\;(\texttt{now}\;d)$ also converges to, and
vice versa.
The forward direction wraps the convergence proof in
\texttt{conv\_later}; the reverse direction inverts one
\texttt{later} step.

\paragraph{The contradiction.}
Any \texttt{CPO.type} in our hierarchy inherits
\texttt{PartialOrder}, which requires Leibniz antisymmetry.
If \texttt{delay}\;$D$ were a \texttt{CPO.type} with an order
extending the convergence preorder, then Fact~C would give
$\texttt{now}\;d \sqsubseteq \texttt{later}\;(\texttt{now}\;d)$
and
$\texttt{later}\;(\texttt{now}\;d) \sqsubseteq \texttt{now}\;d$,
and antisymmetry would force
$\texttt{now}\;d = \texttt{later}\;(\texttt{now}\;d)$---contradicting
Fact~B.

The theorem \texttt{delay\_CPO\_requires\_quotient} is fully proved;
no axioms or admissions are used.
It is a \emph{negative} result: it does not merely say that we have
not found a CPO structure, but that no such structure exists on the
raw \texttt{delay} type under Leibniz equality.

\paragraph{The \texttt{bisim\_trans} guardedness failure.}
The file also develops the monad structure of \texttt{delay} and
attempts to prove that bisimulation is an equivalence relation.
Reflexivity and symmetry are straightforward corecursions.
Transitivity, however, is admitted (\texttt{bisim\_trans}).

The difficulty arises in a single case.
When the first bisimulation proof ends with \texttt{bisim\_right}
(skipping a \texttt{later} on the right) and the second ends with
\texttt{bisim\_left} (skipping a \texttt{later} on the left), the
two steps cancel:
\[
  \underbrace{t_1 \sim \texttt{later}\;s_2}_{\texttt{bisim\_right}}
  \qquad
  \underbrace{\texttt{later}\;s_2 \sim t_3}_{\texttt{bisim\_left}}
\]
The recursive call $\texttt{IH}\;t_1\;s_2\;t_3$ is needed, but it
does not appear under any constructor of \texttt{bisim}---the
\texttt{later} wrappers have been consumed on both sides.
\Rocq{}'s guardedness checker rejects this because the corecursive
call is not productive: it does not produce a constructor before
recurring.

This is a well-known limitation, not specific to our development.
Three resolutions exist in the literature:
\begin{enumerate}
  \item \textbf{Parameterised coinduction} (the \texttt{paco}
    library~\cite{Capretta2005}): replaces the guardedness check
    with an explicit monotonicity argument on a coinduction
    hypothesis, allowing non-guarded recursive calls.
  \item \textbf{Sized types} (available in Agda but not \Rocq{}):
    track the ``depth'' of a coinductive value in its type, making
    the well-foundedness of the recursive call explicit.
  \item \textbf{Mixed inductive-coinductive bisimilarity}
    (Chapman, Uustalu, and Veltri~\cite{ChapmanUustaluVeltri2019}):
    redefine bisimulation with an inductive ``convergence to a
    common value'' component nested inside a coinductive wrapper,
    eliminating the problematic case by construction.
\end{enumerate}
We do not pursue these resolutions because the CPO obstruction
theorem (\texttt{delay\_CPO\_requires\_quotient}) is fully proved
without \texttt{bisim\_trans}, and the flat lift sidesteps the
issue entirely.

\paragraph{The \texttt{converges\_bisim} admission.}
A second result is admitted:
$\texttt{converges\_bisim}$, which states that convergence is
preserved by bisimulation.
The difficulty is analogous: inverting a coinductive bisimulation
hypothesis inside an inductive recursion on convergence leads to
cases where the induction measure does not decrease.
This result is also unused in the obstruction proof and would
be resolved by the same techniques as \texttt{bisim\_trans}.

% ·· Resolution ··············································
\subsection{Resolution: The Flat Lift}
\label{sec:disc-lift-resolution}

The obstruction admits two resolutions, corresponding to two ways of
eliminating the offending pair $\texttt{now}\;d \neq
\texttt{later}\;(\texttt{now}\;d)$.

\paragraph{Quotient by bisimulation.}
The principled resolution is to quotient \texttt{delay}\;$D$ by weak
bisimulation, obtaining a type
$\texttt{delay}\;D\;/\!\sim$ in which
$[\texttt{now}\;d] = [\texttt{later}\;(\texttt{now}\;d)]$ by
construction.
Chapman, Uustalu, and Veltri~\cite{ChapmanUustaluVeltri2019}
carry out this construction in Agda using quotient
inductive-inductive types (QIITs), proving that the quotient admits
a monad structure whose laws hold up to propositional equality.
They further show that the resulting type is isomorphic to the
\emph{partiality monad}---a countable join completion---under the
assumption of countable choice.

\Rocq{} does not natively support quotient types.
Quotients can be simulated using setoids (as BKV does implicitly)
or \texttt{Prop}-truncated sigma types, but both approaches require
carrying setoid morphism proofs through every construction.
The \HB{} hierarchy, designed for Leibniz equality throughout,
would need substantial reworking to accommodate setoid-based types.

\paragraph{The flat alternative.}
The flat lift $D_\bot = \texttt{option}\;D$ avoids the issue entirely
by giving up the step-by-step computational structure.
The bottom element $\bot = \texttt{None}$ and values
$\texttt{Some}\;d$ are propositionally distinct from each other
\emph{and} from $\texttt{Some}\;d'$ when $d \neq d'$, so no
bisimulation quotient is needed.
The cost is the use of \texttt{ClassicalEpsilon} in the supremum
construction (\cref{sec:meth-lift}): the informative excluded middle
is used to decide whether a chain is eventually \texttt{Some}, which
makes the library non-extractable.

\paragraph{Summary of the trade-off.}
\Cref{tab:lift-comparison} summarises the two approaches.
The flat lift fits cleanly into the \HB{} hierarchy with Leibniz
equality; the coinductive lift is computationally more natural but
requires either quotient types or a setoid-based framework.
Our choice of the flat lift is a pragmatic compromise: the library's
purpose is denotational \emph{reasoning} (proving adequacy,
constructing fixed points), not executable extraction, so the
classical axiom is an acceptable price for a simpler infrastructure.

\begin{table}[ht]
\centering
\caption{Comparison of lift constructions.}
\label{tab:lift-comparison}
\small
\begin{tabular}{@{}l l l@{}}
\toprule
\textbf{Aspect} & \textbf{Coinductive lift} & \textbf{Flat lift} \\
\midrule
Type              & \texttt{delay D}
                  & \texttt{option D} \\
Bottom            & $\texttt{cofix}\;\bot := \texttt{later}\;\bot$
                  & \texttt{None} \\
Equality          & Weak bisimulation
                  & Leibniz \texttt{=} \\
Monad laws        & Up to bisimulation
                  & Up to \texttt{=} \\
CPO structure     & Requires quotient types
                  & Direct (\texttt{ClassicalEpsilon}) \\
Extractable       & Yes (in principle)
                  & No (\texttt{ClassicalEpsilon}) \\
\HB{} compatible  & No (without quotients)
                  & Yes \\
\bottomrule
\end{tabular}
\end{table}

\noindent
The development in \texttt{LiftMonad.v} (490 lines) is
self-contained: it imports no files from the library's hierarchy and
could be extracted into a standalone proof of the obstruction
result.
Its purpose is not to provide library infrastructure but to make the
design rationale for the flat lift precise and machine-checked.
