% !TeX root = ../main.tex
% chapters/introduction.tex

\chapter{Introduction}
\label{ch:introduction}

% ── Motivation ────────────────────────────────────────────────────
\section{Motivation}
\label{sec:intro-motivation}

The operational behavior of a program---what it computes on given inputs---can be studied directly by running it. 
But understanding what a program means, in a sense precise enough to support mathematical proof, requires 
something more: a compositional interpretation of the program's syntax in terms of mathematical objects. 
\emph{Denotational semantics} provides this interpretation, assigning to each program a mathematical entity---
its denotation---in a way that respects the structure of the language. For simple programs, denotations 
are straightforward. For programs with recursion and recursive types, however, the construction of a suitable 
mathematical universe is itself a deep problem.

Consider the recursive definition: $f(x) = 1 + f(x)$. As a program, $f$ calls 
itself on every input, computing nothing---it \textit{diverges}. But the program's definition, within this framework of constructing denotations for programs, 
is also a mathematical equation with no solution. Nevertheless, the program is well-formed in any language 
supporting general recursion---so a complete mathematical semantics must assign it a precise meaning. The answer \emph{domain theory} 
gives is elegant: interpret $f$ as the \textit{least fixed point} of the operator $\Phi(g)(x) = 1 + g(x)$, 
computed as the supremum of the sequence $\bot, \Phi(\bot), \Phi^{2}(\bot), \ldots$ in a mathematical structure---an 
$\omega$-complete partial order---where the order is on the amount of information conveyed by the program, and $\bot$, 
being the least element, represents "no information". The operator $\Phi$ is \emph{Scott-continuous}: it preserves the 
order structure in a way that guarantees this sequence has a supremum, and that the supremum is the least fixed point. 
For our example program, that least fixed point is $\bot$ itself, correctly capturing the notion of divergence. This 
result---that every Scott-continuous function on a \emph{pointed} $\omega$-\emph{CPO} has a least fixed point, given 
explicitly as the supremum of its iteration sequence from $\bot$---is the \emph{Kleene fixed-point theorem}, and it 
is the mathematical engine behind the denotational semantics of recursive programs.

Recursive programs, however, are only half of the story. Programming languages also admit recursive \textit{types}:
equations such as $D \cong [D \to D]$, which ask for a mathematical domain isomorphic to its own \textit{continuous}
function space. This equation is not exotic---it is exactly what is needed to model the untyped lambda calculus, where
any term may be applied to any other. No set-theoretic model satisfies it
\footnote{For any set $S$ with more than one element, $\left|[S \to S]\right| = |S|^{|S|} > |S|$ by Cantor's theorem, 
so no set $S$ can satisfy $S \cong [S \to S]$.},
but domain theory does, through a more
sophisticated construction: an inverse limit of an approximating sequence of 
domains connected by structure-preserving maps. Where the Kleene theorem works \textit{inside} a single domain, solving
recursive type equations requires working \textit{across families} of domains. This shift demands that the morphisms 
connecting those families carry their own mathematical structure---specifically, an order---and this is precisely what 
enriched category theory provides. This is the passage from domain theory to enriched domain theory, and it is the central 
organizing principle of this thesis.

The enriched perspective is not merely convenient---it is logically necessary. The inverse-limit construction requires
\textit{embedding-projection} pairs between domains. Informally, such a pair between domains $D$ and $E$ consists of
an "inclusion" map $e : D \to E$ and a "retraction" $p : E \to D$: embedding $D$ inside the larger $E$, 
with $p$ recovering the best approximation in $D$ of any element of $E$. Formally, $e$ and $p$ must satisfy $p \circ e = id_{D}$ 
---the retraction exactly recovers embedded elements---and $e \circ p \sqsubseteq id_{E}$---the composition in the other 
direction is a deflation, approximating but not exceeding the identity. That second condition is an inequality between 
morphisms, which only makes sense when morphisms themselves carry an order structure. But the order on morphisms is not enough on 
its own. For the inverse-limit to exist, composition must interact coherently with that order: given ascending chains of morphisms 
$f_{0} \sqsubseteq f_{1} \sqsubseteq \cdots$ and $g_{0} \sqsubseteq g_{1} \sqsubseteq \cdots$, their compositions must satisfy
$\sqcup_{n}(f_{n} \circ g_{n}) = (\sqcup_{n} \: f_{n}) \circ (\sqcup_{n} \: g_{n})$. This \emph{joint continuity of composition}
---whose very statement requires hom-sets to be ordered and composition to preserve those orders---is precisely the structure of a 
\textit{CPO-enriched category}: a category in which every hom-set is a $\CPO$ and composition is Scott-continuous in each argument. This
is the minimal abstraction in which recursive domain equations can be coherently formulated. We develop this theory formally in 
\cref{ch:background}.

The foundational mechanisation of these ideas is Benton, Kennedy, and Varming's \Rocq{} library \cite{BKV2009}, which formalized
$\omega$-$\CPO$s, the Kleene fixed-point theorem, and gave a complete denotational semantics for PCF$_v$---a call-by-value language with product types, booleans, and general recursion ---including full soundness and computational adequacy. For its stated purpose,
this library was complete. Two subsequent developments, however, create the need for a more expressive foundation. First, richer
type systems---including those with recursive types modeled by equations of the form $D \cong F(D, D)$---require the $\CPO$-enriched 
layer that PCF$_v$ adequacy alone does not demand; the mathematical tools needed simply lay outside the scope of \cite{BKV2009}. Second,
the emergence of quantum programming languages with quantum control flow requires generalizing the order structure of domain theory
in a direction the library was never designed to support. Beyond these theoretical extensions, there is also a practical dimension: 
the Hierarchy Builder framework \cite{HB2023}---now the standard approach for hierarchical mathematical libraries in Rocq---postdates 
\cite{BKV2009} entirely, and the hand-assembled infrastructure of that era makes extending the library difficult in practice. 
This thesis takes \cite{BKV2009} as its classical foundation and extends it in both directions.

The second direction---quantum programming languages---is motivated by a semantic requirement that classical domain theory cannot meet. Most existing quantum languages---
Proto-Quipper, Quipper, and their relatives---support only \textit{classical control flow} over quantum data: the decision of which 
branch to take is always a classical computation on classical information. However, \emph{Quantum control flow}---branching on unmeasured
quantum states---is a strictly more expressive paradigm. The canonical example is the \textit{quantum switch} \cite{CDPV2013}: 
given operations $A$ and $B$ and a control qubit in state $\alpha|0\rangle + \beta|1\rangle$, apply $A$ then $B$ on the $|0\rangle$
branch and $B$ then $A$ on the $|1\rangle$ branch, without measuring the qubit. The result operates coherently on a superposition of
causal orders, a process with no classical analogue, whose indefinite causal order has recently been verified experimentally
\footnote{Richter et al.\ \cite{RichterEtAl2025} report the first device-independent verification protocol: their photonic 
quantum-switch experiment violates a Bell-like causal inequality by 18 standard deviations, demonstrating correlations 
incompatible with any definite ordering of events.}. Classical domain theory cannot model this: its order structure $x \sqsubseteq y$ is 
Boolean-valued---a proposition that is simply true or false---and cannot itself carry quantum information or exist in superposition. 
Modeling quantum-coherent ordering requires a richer structure. The natural replacement is a \emph{quantale-valued} order $R(x, y) \in Q$, 
where $Q$ is an \textit{involutive quantale}---a noncommutative lattice whose elements carry quantum information, defined precisely 
in \cref{sec:disc-qcpo}. This yields \emph{quantum CPOs} ($\qCPO$s), introduced by Kornell, Lindenhovius, and Mislove \cite{KLM2024}, who 
show that the resulting category is algebraically compact in the relevant sense: recursive type equations have canonical solutions, 
providing the semantic foundations needed for quantum type systems with recursion. The crucial observation for this thesis is structural. 
Given quantum $\CPO$s $X$ and $Y$, the quantum Scott-continuous maps $[X \to_{q} Y]$ form a $\CPO$ under the pointwise quantum order. 
That is, $\qCPO$ hom-sets are themselves $\CPO$s---precisely the definition of a $\CPO$-enriched category. The same 
abstract enriched framework developed for classical recursive types therefore applies to quantum domains without 
modification.

The choice to develop this theory as a mechanized library reflects the particular demands of the subject. \emph{Proof assistants} reject
informal argument: every definition must be fully specified and every proof must close under \emph{kernel} verification. In a subject 
as technically intricate as domain theory---where subtle conditions on order, continuity, and chain stability interact throughout
---this discipline is not merely convenient but epistemically significant. Mechanisation also enables a qualitatively different form 
of reuse: once $\qCPO$ is registered as a \textbf{CPOEnrichedCat} instance, \Rocq{}'s elaborator automatically locates and applies every
enriched theorem to quantum domain contexts, with machine-checked evidence that every implicit hypotheses is satisfied and no condition
has been silently assumed --- unlike citing a paper theorem. Finally, the Hierarchy Builder framework makes the deep dependency structure 
of domain theory---where each algebraic layer extends the last---manageable in a way that hand-assembled infrastructure cannot sustain 
at scale.

This thesis presents a unified, modern \Rocq{} library formalizing this perspective, taking \cite{BKV2009} as its classical 
starting point and building outward. The library comprises three layers: a modernized classical development (partial orders through the 
Kleene fixed-point theorem, the lift monad, and a complete PCF$_v$ adequacy proof), a $\CPO$-enriched category formalization 
(enriched categories, locally continuous functors, natural transformations, and the framework for recursive domain equations via 
EP-pairs and bilimits), and a quantum extension that demonstrates $\qCPO$ satisfies the axioms of a 
\textbf{CPOEnrichedCat}---making the full enriched toolkit applicable to quantum domains. The detailed contributions are enumerated in 
\cref{sec:intro-contributions}.


% NOTE: You can cross-reference later chapters like this:
%   \cref{ch:background}  →  "Chapter 2"
%   \cref{sec:bg-cpo}     →  "Section 2.1.2"
% cleveref handles the "Chapter"/"Section" prefix automatically.

% ── Contributions ─────────────────────────────────────────────────
\section{Contributions}
\label{sec:intro-contributions}

The contributions of this thesis are as follows.

\begin{enumerate}

\item \textbf{A modernized classical domain theory library.}
  We develop approximately 15{,}000 lines of \Rocq{} 9.0, built on the Hierarchy Builder framework 
  \cite{HB2023}. The library provides an HB hierarchy from preorders through pointed $\omega$-$\CPO$s, 
  with concrete instances for products, sums, function spaces, and the lift monad. We prove the 
  \emph{Kleene fixed-point theorem} and register the fixed-point operator $\fixp$ as a Scott-continuous 
  map on the function space $[D \to_{c} D] \to_{c} D$, making it available as a first-class object in 
  further constructions. We also formalize the Scott topology and prove the equivalence between the 
  computational and topological notions of continuity. Every classical construction in \cite{BKV2009} 
  is reproduced, and the HB infrastructure reduces defining a new CPO instance to a single declaration.

\item \textbf{A $\CPO$-enriched category formalization.}
  We formalize $\CPO$-enriched categories---categories whose hom-sets are $\CPO$s and whose composition 
  is jointly Scott-continuous---together with locally continuous functors and enriched natural 
  transformations. For natural transformations we prove vertical composition, left and right whiskering, 
  and the existence of pointwise suprema for chains of natural transformations. As a validation, we prove 
  an enriched Yoneda lemma, establishing a $\CPO$ isomorphism between natural transformations out of a 
  representable functor and the representing object. This layer is entirely original: \cite{BKV2009} had no 
  enriched-categorical structure, and while recent work has formalized general V-enrichment in a univalent 
  setting \cite{VanDerWeide2024}, no prior library targets $\CPO$-enrichment for domain-theoretic applications.

\item \textbf{Recursive domain equations via EP-pairs and bilimits.}
  Building on the enriched layer, we formalize \textit{embedding-projection pairs}, their composition 
  and deflation properties, and \textit{mixed-variance locally continuous bifunctors}. We construct the 
  approximating sequence for a recursive domain equation $D \cong F(D, D)$ and define the 
  \texttt{BilimitData} specification that packages a solution: a limit object together with 
  \texttt{ROLL} and \texttt{UNROLL} isomorphisms satisfying the required coherence conditions. 
  The construction isolates a single axiom---the existence of the bilimit---whose proof obligation 
  is explicit and whose consequences (the full isomorphism and the supremum-of-deflations identity) 
  are derived formally.

\item \textbf{A complete PCF$_v$ adequacy proof.}
  We give a full denotational semantics for PCF$_v$---a call-by-value variant of PCF with natural 
  numbers, booleans, products, and general recursion---and prove both soundness and computational 
  adequacy. The syntax uses intrinsically typed de~Bruijn indices with a computational case-analysis 
  combinator (\texttt{var\_case}) that avoids the \texttt{JMeq} axiom, ensuring that substitution 
  lemmas hold definitionally. Adequacy is established via a type-indexed logical relation: we prove 
  that every well-typed term satisfies the relation (the fundamental lemma), and derive as a corollary 
  that a closed expression converges operationally if and only if its denotation is defined. 
  The proof follows the structure of \cite{BKV2009} but is carried out entirely within the 
  HB-based CPO hierarchy, demonstrating that the modernized infrastructure supports a 
  complete language formalization.

\item \textbf{Quantum $\CPO$ formalization and the enrichment correspondence.}
  Following Kornell, Lindenhovius, and Mislove \cite{KLM2024}, we formalize involutive quantales, 
  quantum posets, and quantum $\CPO$s, including the quantum Kleene fixed-point theorem. We then 
  establish that the quantum Scott-continuous maps $[X \to_{q} Y]$ between quantum $\CPO$s form a 
  classical $\CPO$ under the pointwise quantum order. This is the thesis's central structural claim: 
  both $\CPO$ and $\qCPO$ satisfy the axioms of a $\CPO$-enriched category, and consequently the 
  entire enriched framework developed in contributions (2) and (3)---locally continuous functors, 
  natural transformations, and the machinery for recursive domain equations---applies to quantum 
  domains without modification.

\end{enumerate}


% ── Thesis outline ────────────────────────────────────────────────
\section{Thesis Outline}
\label{sec:intro-outline}

% TODO: Brief chapter-by-chapter summary
