% !TeX root = ../main.tex
% chapters/methods.tex

\chapter{Methods}
\label{ch:methods}

% ── Library architecture ──────────────────────────────────────────
\section{Library Architecture}
\label{sec:meth-architecture}

The library is organised into five directories, each compiled as a separate
Rocq theory by the Dune build system.  The dependency graph is strictly
layered: each directory may import from the directories above it but never
in the reverse direction.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  dir/.style={draw, rounded corners, minimum width=3.6cm, minimum height=0.9cm,
              font=\sffamily\small, fill=gray!8},
  arr/.style={-{Stealth[length=6pt]}, thick},
  plan/.style={-{Stealth[length=6pt]}, thick, dashed, gray},
  node distance=1.3cm and 2.4cm,
]
  \node[dir] (str) {\texttt{structures/}};
  \node[dir, below=of str] (thy) {\texttt{theory/}};
  \node[dir, below=of thy] (inst) {\texttt{instances/}};
  \node[dir, below left=of inst] (lang) {\texttt{lang/}};
  \node[dir, below right=of inst] (qua) {\texttt{quantum/}};

  \draw[arr] (thy) -- (str);
  \draw[arr] (inst) -- (thy);
  \draw[arr] (lang) -- (inst);
  \draw[arr] (qua) -- (inst);
  \draw[plan] (lang) -- node[below, font=\scriptsize\itshape, text=gray]
    {planned} (qua);
\end{tikzpicture}
\caption{Directory dependency tree.  Solid arrows denote actual imports;
  the dashed arrow marks a dependency that the Dune build files pre-declare
  for the planned \texttt{QMiniCore} quantum language but that no current
  source file uses.  Dependencies are transitive: \texttt{lang/} may
  import from any directory above it.}
\label{fig:dep-graph}
\end{figure}

The five layers serve distinct roles:
\begin{itemize}
  \item \texttt{structures/}\enspace contains the HB mixin and structure
    declarations that define the algebraic hierarchy---partial orders,
    $\omega$-CPOs, pointed CPOs, and CPO-enriched categories.  These files
    declare \emph{what} the structures are but prove nothing beyond the most
    immediate consequences of the axioms.

  \item \texttt{theory/}\enspace contains derived mathematics: the
    chain-stabilisation lemmas, fixed-point theory, the Scott topology, the
    CPO constructions (products, sums, lifting, function spaces), enriched
    natural transformations, and the domain-equation machinery.

  \item \texttt{instances/}\enspace registers concrete models as HB
    instances.  In particular, \texttt{Function.v} registers the function-space
    CPO $[D \to_c E]$ as an instance of \texttt{CPOEnrichedCat}---the
    self-enrichment of \textbf{CPO}---and \texttt{FunLift.v} registers the
    mixed-variance bifunctor $D \mapsto [D \to_c D_\bot]$ as a
    \texttt{MixedLCFunctor} instance.

  \item \texttt{quantum/}\enspace formalises the quantum domain theory of
    Kornell, Lindenhovius, and Mislove: involutive quantales, quantum posets,
    quantum CPOs, and quantum-continuous morphisms.

  \item \texttt{lang/}\enspace formalises the PCF$_v$ language---its syntax,
    operational semantics, denotational semantics, soundness, and adequacy---and
    will host the quantum language \texttt{QMiniCore} when it is specified.
\end{itemize}

This layering enforces a clean separation of concerns.  A file in
\texttt{structures/} cannot depend on a proof in \texttt{theory/}; a file
in \texttt{theory/} cannot depend on a concrete instance.  Several of the
design decisions discussed in \cref{sec:meth-design-decisions} are
direct consequences of maintaining this discipline---in particular, the
separate-continuity axioms for enriched composition (\cref{sec:meth-enriched-structure})
exist precisely to avoid a circular dependency between \texttt{structures/}
and \texttt{theory/}.


% ── The algebraic hierarchy ───────────────────────────────────────
\subsection{The Algebraic Hierarchy}
\label{sec:meth-hierarchy}

The library's type hierarchy is declared using Hierarchy Builder.  As
described in \cref{sec:bg-hb}, HB synthesises bundled structures,
coercion paths, and canonical projections from user-written mixins.
The hierarchy splits into two tracks: an \emph{order track} for domain
theory and an \emph{enriched track} for enriched category theory.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  mixin/.style={draw, rounded corners=3pt, minimum width=2.4cm,
                minimum height=0.7cm, font=\small\ttfamily, fill=blue!6},
  struct/.style={draw, rounded corners=3pt, minimum width=2.4cm,
                 minimum height=0.7cm, font=\small\sffamily\bfseries, fill=orange!10},
  arr/.style={-{Stealth[length=5pt]}, semithick},
  node distance=0.85cm and 1.8cm,
]
  % ── Order track ──
  \node[mixin] (hasle) {HasLe};
  \node[mixin, below=of hasle] (ispre) {IsPreorder};
  \node[mixin, below=of ispre] (ispo) {IsPartialOrder};
  \node[mixin, below=of ispo] (hassup) {HasSup};
  \node[mixin, below=of hassup] (iscpo) {IsCPO};

  % Pointed branch
  \node[mixin, below left=1.0cm and 1.6cm of iscpo] (hasbot) {HasBottom};
  \node[mixin, below=of hasbot] (isptd) {IsPointed};

  % Bundled structures (right)
  \node[struct, right=3.2cm of ispre] (pre) {Preorder};
  \node[struct, right=3.2cm of ispo] (po) {PartialOrder};
  \node[struct, right=3.2cm of iscpo] (cpo) {CPO};
  \node[struct, right=3.2cm of isptd] (pcpo) {PointedCPO};

  % ── Enriched track ──
  \node[mixin, below right=1.0cm and 1.6cm of iscpo] (hashom) {HasHom};
  \node[mixin, below=of hashom] (hasid) {HasId};
  \node[mixin, below=of hasid] (iscpoe) {IsCPOEnriched};

  \node[struct, right=3.2cm of iscpoe] (cpoecat) {CPOEnrichedCat};

  % ── Arrows (order track) ──
  \draw[arr] (ispre) -- (hasle);
  \draw[arr] (ispo) -- (ispre);
  \draw[arr] (hassup) -- (ispo);
  \draw[arr] (iscpo) -- (hassup);
  \draw[arr] (hasbot) -- (iscpo);
  \draw[arr] (isptd) -- (hasbot);

  % ── Arrows (enriched track) ──
  \draw[arr] (hashom) -- (iscpo);
  \draw[arr] (hasid) -- (hashom);
  \draw[arr] (iscpoe) -- (hasid);

  % ── Arrows to bundled structures ──
  \draw[arr, dashed, gray] (ispre) -- (pre);
  \draw[arr, dashed, gray] (ispo) -- (po);
  \draw[arr, dashed, gray] (iscpo) -- (cpo);
  \draw[arr, dashed, gray] (isptd) -- (pcpo);
  \draw[arr, dashed, gray] (iscpoe) -- (cpoecat);
\end{tikzpicture}
\caption{The HB mixin hierarchy.  Blue boxes are mixins (bundles of data
  or axioms); orange boxes are the bundled structures that HB synthesises.
  Solid arrows indicate mixin dependencies; dashed arrows indicate which
  mixins contribute to each structure.}
\label{fig:mixin-hierarchy}
\end{figure}

The order track begins with \texttt{HasLe}, which attaches a binary
relation~$\sqsubseteq$ to a type.  Successive mixins add reflexivity
and transitivity (\texttt{IsPreorder}), antisymmetry
(\texttt{IsPartialOrder}), a supremum function (\texttt{HasSup}), and the
least-upper-bound axioms (\texttt{IsCPO}).  A parallel branch adds a
bottom element (\texttt{HasBottom}) and its axiom (\texttt{IsPointed}).
Each of these follows the \emph{two-mixin idiom}: one mixin provides the
data (a function or an element) and a second mixin provides the
specification (the axioms that the data must satisfy).  For example:

\begin{lstlisting}[caption={The two-mixin idiom for $\omega$-CPOs.},
                    label={lst:iscpo}]
HB.mixin Record HasSup T of PartialOrder T :=
  { sup : chain T → T }.

HB.mixin Record IsCPO T of HasSup T & PartialOrder T := {
  sup_upper : ∀ (c : chain T) (n : nat), c.[n] ⊑ ⊔ c;
  sup_least : ∀ (c : chain T) (x : T),
                (∀ n, c.[n] ⊑ x) → ⊔ c ⊑ x;
}.
\end{lstlisting}

\noindent
This separation is idiomatic in HB-based libraries: the data mixin can
be instantiated mechanically (often by providing a single function), while
the axiom mixin is where the mathematical content lives.  HB then
synthesises the bundled structure \texttt{CPO} from the accumulated mixins,
together with coercion paths that allow any \texttt{CPO.type} to be used
wherever a \texttt{PartialOrder.type} or \texttt{Preorder.type} is expected.

The enriched track extends the hierarchy into enriched category theory.
\texttt{HasHom} equips a type of objects with a family of hom-CPOs,
\texttt{HasId} adds identity morphisms, and \texttt{IsCPOEnriched} adds
composition together with its categorical laws and Scott-continuity axioms.
This track is discussed in detail in \cref{sec:meth-enriched}.

\paragraph{Plain records.}
Not every type in the library is an HB structure.  The types
$\omega$-chains, monotone functions, and Scott-continuous functions are
defined as plain \texttt{Record}s parameterised over HB-structured
carriers:

\begin{lstlisting}[caption={The \texttt{cont\_fun} record.},
                    label={lst:contfun}]
Record cont_fun (D E : CPO.type) : Type := {
  cf_mono :> mono_fun D E;
  cf_cont : continuous cf_mono;
}.
\end{lstlisting}

\noindent
A \texttt{cont\_fun} bundles a monotone function with a proof of
Scott-continuity.  The coercion \lstinline{:>} on \texttt{cf\_mono}
allows a continuous function to be used wherever a monotone function or
a plain function is expected, via the chain
$\texttt{cont\_fun} \to \texttt{mono\_fun} \to (D \to E)$.
These types are not declared as HB structures because they are
\emph{parameterised} by their domain and codomain types, whereas HB
structures are indexed by a single carrier.  Migrating them would require
converting every constructor call site (over fifty across twelve files)
for minimal practical gain, since the existing identity and composition
combinators (\texttt{cont\_id}, \texttt{cont\_comp}) already provide the
categorical infrastructure needed.


% ── Design decisions ──────────────────────────────────────────────
\subsection{Design Decisions}
\label{sec:meth-design-decisions}

Several non-obvious architectural choices arose during the modernisation
of the Benton--Kennedy library for Rocq~9.0 and HB.  This section records
the decisions that affect the global structure of the library; further
decisions specific to individual modules are discussed where they arise
(e.g., the separate-continuity axioms in \cref{sec:meth-enriched-structure},
the \texttt{var\_case} combinator in \cref{sec:meth-pcf-syntax}).

\paragraph{Organise by dependency, not by concept (DD-001).}
An earlier version of the codebase had a separate \texttt{Pointed.v}
file, following a one-file-per-concept organisation in which pointed
CPOs, bottom elements, and strict functions each had their own module.
During the HB restructuring, this was replaced by a dependency-driven
layout: \texttt{HasBottom} and \texttt{IsPointed} live in
\texttt{CPO.v} (because \texttt{IsPointed} depends on \texttt{CPO~T}),
and \texttt{strict\_fun} lives in \texttt{Morphisms.v} (because it
depends on \texttt{cont\_fun}).  This follows the MathComp convention:
each definition is placed in the file determined by its hard
dependencies, not by an idealised conceptual grouping.

\paragraph{Monotonicity before continuity in mixins (DD-003).}
In every mixin where both monotonicity and Scott-continuity are required
(e.g., \texttt{IsCPOEnriched}, \texttt{IsLocallyContinuous}), monotonicity
fields are declared \emph{before} continuity fields.  The \texttt{continuous}
predicate takes a \texttt{mono\_fun} as its argument, so a continuity
field that refers to \texttt{Build\_mono\_fun~\ldots~comp\_mono\_l}
requires \texttt{comp\_mono\_l} to already be in scope.  If the order
is reversed, the mixin fails to elaborate.  This is a Rocq--HB
implementation constraint, not a mathematical preference.

\paragraph{Leibniz equality for categorical laws (DD-004).}
The categorical laws in \texttt{IsCPOEnriched}---identity, associativity,
and the unit laws---use Leibniz equality~($=$) rather than a setoid
relation~($\approx$).  The hom-CPOs in our framework are
\texttt{CPO.type} values, and equality on their elements is Leibniz
equality by default.  This keeps the laws definitionally simple and avoids
the overhead of setoid-rewriting infrastructure at the structure level.
If a concrete instance used a coarser extensional equality, the laws
would need to be restated in the relevant theory file.

\paragraph{Plain records instead of universe polymorphism (DD-019).}
Registering HB instances for functors and natural transformations on
\texttt{CPOEnrichedCat.type} triggers Rocq universe inconsistencies:
the quantification over all objects~$X : \mathcal{C}$ in a natural
transformation forces the carrier universe of the functor to be at
least as large as the universe of~$\mathcal{C}$, conflicting with HB's
constraint that structure carriers inhabit a fixed universe level.
Rocq's \texttt{Set Universe Polymorphism} could in principle resolve
this by parameterising each definition over its universe level, but
adopting it would require pervasive annotations across the entire HB
hierarchy---every structure, mixin, and instance in the library would
need explicit universe declarations---for no practical benefit beyond
the enriched layer.  Instead, the library represents locally continuous
functors (\texttt{lc\_functor}) and natural transformations
(\texttt{nat\_trans}) as plain Rocq records outside HB, sidestepping the
universe machinery entirely.  A conversion function
\texttt{lc\_functor\_of\_hb} bridges from an HB
\texttt{LocallyContinuousFunctor} instance to the plain record
when needed.

\paragraph{Instance ordering for \texttt{PointedCPO} (DD-008).}
When registering a concrete type as a \texttt{PointedCPO}, the three HB
instances---\texttt{HasBottom}, \texttt{IsCPO}, \texttt{IsPointed}---must
be registered in exactly this order.  HB checks the full diamond of mixin
dependencies at each registration: \texttt{HasBottom} requires
\texttt{HasLe} (which is a prerequisite of \texttt{IsCPO}), and
\texttt{IsPointed} requires both \texttt{CPO} and \texttt{HasBottom}.
Registering in the wrong order causes a \emph{silent} elaboration failure:
the instance simply does not register, with no error message.  This was
the source of a bug in an early draft of \texttt{Lift.v} and is now a
documented invariant.

% ── Constructions ─────────────────────────────────────────────────
\section{CPO Constructions}
\label{sec:meth-constructions}

With the algebraic hierarchy in place, the library proceeds to the
standard domain-theoretic toolkit: products, sums, lifting, and function
spaces.  Each construction equips a new type with HB instances, making
it available as a first-class CPO for all downstream development.  The
mathematical definitions were given in \cref{ch:background}; this
section focuses on how they are realised in Rocq and the engineering
choices that arise.


% ·· Products and Sums ····················································
\subsection{Products and Sums}
\label{sec:meth-products-sums}

% Source: theory/Products.v (~535 lines), theory/Sums.v (~625 lines)

\paragraph{Products.}
The product CPO $D \times E$ uses Rocq's built-in \texttt{prod} type
with the componentwise order:
$(d_1, e_1) \sqsubseteq (d_2, e_2)$ iff
$d_1 \sqsubseteq d_2$ and $e_1 \sqsubseteq e_2$.
The supremum of a chain~$c$ is thus defined componentwise:
$\bigsqcup c = (\bigsqcup \pi_1(c),\; \bigsqcup \pi_2(c))$,
where $\pi_1(c)$ and $\pi_2(c)$ are the chains obtained by applying
the (monotone) projections pointwise.  The file registers HB instances
for \texttt{PartialOrder}, \texttt{CPO}, and---when $D$ and $E$ are
pointed---\texttt{PointedCPO} on \texttt{D~*~E}, with
$\bot_{D \times E} = (\bot_D, \bot_E)$.

The projections $\pi_1$, $\pi_2$ and the pairing combinator
$\langle f_1, f_2 \rangle$ are packaged as \texttt{cont\_fun} values.
The full universal property of categorical products is then proved:
$\beta$-reduction ($\pi_i \circ \langle f_1, f_2 \rangle = f_i$),
$\eta$-expansion ($h = \langle \pi_1 \circ h,\; \pi_2 \circ h \rangle$),
uniqueness, functoriality ($\langle f \circ k,\; g \circ k \rangle
= \langle f, g \rangle \circ k$), and the identity law
($\langle \pi_1, \pi_2 \rangle = \mathrm{id}$).
Continuity of the projections holds by definitional computation: the
proof term is \texttt{reflexivity}.

\paragraph{Sums.}
The separated sum $D + E$ uses Rocq's built-in \texttt{sum} type.
Two elements are comparable only when they belong to the same summand:
\begin{lstlisting}[caption={The separated sum order.},
                    label={lst:sumle}]
Definition sum_le (x y : D + E) : Prop :=
  match x, y with
  | inl a, inl a' => a ⊑ a'
  | inr b, inr b' => b ⊑ b'
  | _,     _      => False
  end.
\end{lstlisting}

\noindent
Because the order never allows crossing from \texttt{inl} to
\texttt{inr}, every $\omega$-chain in $D + E$ is
\emph{summand-stable from index~0}: if $c_0 = \texttt{inl}\;d$ then
$c_n = \texttt{inl}\;a_n$ for all~$n$.  The proof is a one-step
argument by monotonicity---if $c_n$ were in the other summand, we would
have $c_0 \sqsubseteq c_n$ crossing summands, which is~\texttt{False}.
No classical axiom is needed.

Stability lets us define the supremum by matching on $c_0$: extract the
inner chain in the relevant summand, take its supremum, and re-inject.
Extraction uses a total function with a default value for the
``wrong'' summand, avoiding dependent types in the chain machinery.

The injections $\iota_1$, $\iota_2$ and the copairing
$[f, g]$ (\texttt{sum\_case}) are proved continuous.  The full universal
property of coproducts holds:
$\beta$-reduction ($[f, g] \circ \iota_i = f_i$),
$\eta$-expansion ($h = [h \circ \iota_1,\; h \circ \iota_2]$),
uniqueness, and the fusion law
$h \circ [f, g] = [h \circ f,\; h \circ g]$.

An important structural observation: the separated sum is \emph{not}
pointed even when $D$ and $E$ are, because
$\texttt{inl}\;\bot_D$ and $\texttt{inr}\;\bot_E$ are incomparable.
To obtain a pointed coproduct, one applies the lift: $(D + E)_\bot$.


% ·· Lifting ··············································
\subsection{Lifting}
\label{sec:meth-lift}

% Source: theory/Lift.v (~635 lines)

The lift construction~$D_\bot$ adjoins a fresh bottom element to any
CPO, modeling potential non-termination.  The carrier is Rocq's
\texttt{option~D}, with the flat extension of the order:
$\texttt{None} \sqsubseteq x$ for all~$x$,
\;$\texttt{Some}\;d \sqsubseteq \texttt{Some}\;d'$ iff
$d \sqsubseteq d'$,
\;and $\texttt{Some}\;d \sqsubseteq \texttt{None}$ is~\texttt{False}.

\paragraph{Chain stability and the supremum.}
Chains in $D_\bot$ enjoy a stability property analogous to the sum case:
once a chain reaches \texttt{Some}, it can never return to
\texttt{None}.  This observation drives the supremum construction:

\begin{lstlisting}[caption={The lift supremum.},
                    label={lst:liftsup}]
Definition lift_sup (c : chain (option D)) : option D :=
  match excluded_middle_informative (exists n, c.[n] <> None) with
  | right _   => None
  | left  Hex =>
    let '(exist _ N HN) :=
        constructive_indefinite_description _ Hex in
    let d0 := extract_some (c.[N]) HN in
    Some (⊔ D_chain N d0 c HN)
  end.
\end{lstlisting}

\noindent
If every element of the chain is \texttt{None}, the supremum is
\texttt{None}.  Otherwise, \texttt{constructive\_indefinite\_description}
(from the \texttt{ClassicalEpsilon} axiom) extracts a witness
index~$N$ at which $c_N \neq \texttt{None}$.  From~$N$ onward, the
chain lives entirely in~\texttt{Some}, so we extract the inner
$D$-values into a chain on~$D$, take its supremum, and wrap the result
in~\texttt{Some}.

This is the only use of \texttt{ClassicalEpsilon} in the core library
outside of \texttt{ScottTopology.v}.  It provides
\texttt{excluded\_middle\_informative} (decidable excluded middle on
\texttt{Prop}) together with \texttt{constructive\_indefinite\_description}
(witness extraction from existentials).  The factored computation lemma
\texttt{lift\_sup\_some\_eq}---stating that for any witness~$N$ and
default value~$d_0$, $\bigsqcup c = \texttt{Some}(\bigsqcup
D_{\mathrm{chain}})$---serves as the central tool for both the
\texttt{ret} and \texttt{kleisli} continuity proofs.

\paragraph{Monadic structure.}
The lift carries a monad structure in the Kleisli
formulation~(\cref{def:lift}):
%
\begin{align*}
  \mathit{ret}(d) &= \texttt{Some}(d) \\
  \mathit{kleisli}(f)(\texttt{None}) &= \texttt{None} \\
  \mathit{kleisli}(f)(\texttt{Some}\;d) &= f(d)
\end{align*}
%
Both are packaged as \texttt{cont\_fun} values.  Continuity of
\texttt{ret} follows because the mapped chain
$n \mapsto \texttt{Some}(c_n)$ is always in~\texttt{Some}, so
\texttt{lift\_sup\_some\_eq} applies with $N = 0$.  Continuity of
\texttt{kleisli}~$f$ splits into two cases: if the input chain is
everywhere~\texttt{None}, both sides are~\texttt{None}; otherwise,
\texttt{lift\_sup\_some\_eq} reduces the problem to the continuity
of~$f$ on the extracted $D$-chain, with an index-shift argument
aligning the mapped chains.

The three monad laws---left unit, right unit, and associativity---all
reduce to \texttt{cont\_fun\_ext} followed by case analysis on
\texttt{option~D}; each branch closes by \texttt{reflexivity}.
The trade-off between this classical, extensional lift and the
coinductive delay monad---which was used in the original
Benton--Kennedy--Varming library~\cite{BKV2009}---is discussed in
\cref{ch:discussion}.


% ·· Function Spaces ·····································
\subsection{Function Spaces}
\label{sec:meth-function-spaces}

% Source: theory/FunctionSpaces.v (~878 lines)

The constructions so far---products, sums, and the lift---give us the
basic type constructors and the means to model partiality.  What they do
not yet provide is a way to interpret \emph{function types}: to give
meaning to $\tau_1 \to \tau_2$ we need the collection of continuous maps
$[\sem{\tau_1} \to_c \sem{\tau_2}_\bot]$ to be an $\omega$-CPO in its
own right.  This is the function-space construction, and the fact that
it produces a well-behaved exponential object is what makes $\CPO$
cartesian closed (\cref{sec:bg-ccc}).  The formalisation in
\texttt{FunctionSpaces.v} establishes this structure.

The function space $[D \to_c E]$ is the type of Scott-continuous
functions from~$D$ to~$E$, ordered pointwise:
$f \sqsubseteq g$ iff $f(x) \sqsubseteq g(x)$ for all~$x$.
The supremum of a chain of continuous functions is again taken pointwise:
$(\bigsqcup_n f_n)(x) = \bigsqcup_n f_n(x)$.  Proving that this
pointwise supremum is itself Scott-continuous is the central obligation;
it requires a double-indexed argument showing that the family of chains
$\{n \mapsto f_n(c_k)\}_{k}$ is coherent.  When $E$ is pointed, the
constant-$\bot$ function serves as the bottom element, yielding a
\texttt{PointedCPO} instance.

\paragraph{Cartesian closed structure.}
The three continuous maps that witness cartesian closedness
(\cref{prop:cpo-ccc}) are formalised as follows:
\begin{itemize}
  \item $\mathit{eval} : [D \to_c E] \times D \to E$, defined by
    $\mathit{eval}(f, d) = f(d)$.  Continuity requires a diagonal
    argument: for chains $f_n$ and $d_n$, we have
    $f_n(d_k) \sqsubseteq f_{\max(n,k)}(d_{\max(n,k)}) = \mathit{eval}(f_{\max(n,k)}, d_{\max(n,k)})$.
  \item $\mathit{curry}(f) : C \to [D \to_c E]$, defined by
    $\mathit{curry}(f)(c)(d) = f(c, d)$.
  \item $\mathit{uncurry}(g) : C \times D \to E$, defined as
    $\mathit{eval} \circ (g \times \mathrm{id})$.
\end{itemize}
The exponential adjunction
$[C \times D \to_c E] \cong [C \to_c [D \to_c E]]$
is witnessed by \texttt{curry\_uncurry} and \texttt{uncurry\_curry},
both of which hold by \texttt{reflexivity} on the underlying function
equality.  The eval-curry equation from
\cref{fig:exponential-transpose} is likewise definitional:

\begin{lstlisting}[
  caption={Exponential adjunction and the eval-curry law
    (\texttt{FunctionSpaces.v}).  Every proof reduces to function
    extensionality followed by \texttt{reflexivity}, showing that the
    CCC structure is \emph{computational}: the isomorphism holds by
    $\iota\beta$-reduction alone.},
  label={lst:exp-adjunction}
]
Lemma curry_uncurry (f : [C →c [D →c E]]) :
  cont_curry (cont_uncurry f) = f.
Proof.
  apply cont_fun_ext; intros c.
  apply cont_fun_ext; intros d.
  reflexivity.
Qed.

Lemma uncurry_curry (f : [C * D →c E]) :
  cont_uncurry (cont_curry f) = f.
Proof.
  apply cont_fun_ext; intros [c d].
  reflexivity.
Qed.

Lemma eval_curry (f : [C * D →c E]) :
  cont_comp cont_eval
    (cont_prod_map (cont_curry f) (cont_id D)) = f.
Proof.
  apply cont_fun_ext; intros [c d].
  reflexivity.
Qed.
\end{lstlisting}

\paragraph{The continuous fixed-point operator.}
Background \cref{sec:bg-fixedpoints} constructs the least fixed point
$\fixp(f) = \bigsqcup_n f^n(\bot)$ for a single continuous
endomorphism.  To interpret the \textbf{fix} combinator of PCF$_v$
compositionally, however, we need $\fixp$ to be a \emph{morphism} in
$\CPO$---an element of the exponential object
$[[D \to_c D] \to_c D]$---so that it can be composed with the other
semantic maps via the CCC structure of \cref{sec:bg-ccc}.

The formalisation packages this as
$\mathrm{FIXP} : [[D \to_c D] \to_c D]$.
The key continuity obligation is showing that
$\bigsqcup_n \mathrm{fixp}(f_n)$ is a pre-fixed point of
$\bigsqcup_n f_n$.  A secondary complication is that $\fixp$ requires
a \emph{pointed} CPO (for the base element~$\bot$), while the
function-space construction produces an ordinary CPO; coercion lemmas
bridge the gap between \texttt{PointedCPO.type} and
\texttt{CPO.type}.

\texttt{FIXP} satisfies the expected interface: the computation rule
$\mathrm{FIXP}(f) = f(\mathrm{FIXP}(f))$
(\texttt{FIXP\_is\_fixedpoint}), the minimality property
$f(x) \sqsubseteq x \Rightarrow \mathrm{FIXP}(f) \sqsubseteq x$
(\texttt{FIXP\_least}), and fixed-point induction
(\texttt{FIXP\_ind}).  The last of these is the main proof tool for the
adequacy theorem in \cref{ch:results}: to show that a property~$P$
holds of~$\mathrm{FIXP}(f)$, it suffices to show that $P$ is
admissible (\cref{def:admissible}), that $P(\bot)$ holds, and that $P$
is preserved by~$f$.

% ── Enriched category formalization ───────────────────────────────
\section{CPO-Enriched Categories}
\label{sec:meth-enriched}

% Source: structures/Enriched.v, theory/EnrichedTheory.v

The CPO constructions of \cref{sec:meth-constructions} give us products,
sums, lifting, and function spaces, all as $\omega$-CPOs with the
expected universal properties.  These constructions are the raw
material of a denotational semantics, but using them compositionally
requires a categorical framework: one in which composition,
functorial actions, and natural transformations are all well-behaved
with respect to the CPO structure.  That framework is CPO-enrichment
(\cref{sec:bg-enriched}).  This section describes how the enriched
categorical layer is formalised.

% ·· The enriched category structure ························
\subsection{The Enriched Category Structure}
\label{sec:meth-enriched-structure}

% Source: structures/Enriched.v (~388 lines)

The enriched categorical structure is declared in \texttt{Enriched.v}
using the same two-mixin HB idiom as the CPO hierarchy.  The data is
introduced in three stages:

\begin{enumerate}
  \item \texttt{HasHom}~equips a type \texttt{Obj} of objects with a
    family of hom-CPOs: for every pair \texttt{A~B~:~Obj}, the term
    \texttt{hom~A~B} has type \texttt{CPO.type}.
  \item \texttt{HasId}~provides an identity morphism
    \texttt{id\_mor~A~:~hom~A~A} for each object.
  \item \texttt{IsCPOEnriched}~bundles composition, its continuity
    witnesses, and the categorical laws into a single mixin.
\end{enumerate}

\begin{lstlisting}[
  caption={The \texttt{IsCPOEnriched} mixin (\texttt{Enriched.v}).
    Composition is declared inside the mixin rather than in a separate
    \texttt{HasComp} mixin because HB's global projection for
    \texttt{comp} requires canonical-structure resolution on the
    carrier type, which is unavailable at the mixin stage.},
  label={lst:iscpoenriched}
]
HB.mixin Record IsCPOEnriched Obj
    of HasHom Obj & HasId Obj := {
  comp : forall {A B C : Obj},
    hom B C -> hom A B -> hom A C;
  comp_mono_l : forall {A B C} (f : hom B C),
    monotone (hom A B) (hom A C) (fun g => comp f g);
  comp_mono_r : forall {A B C} (g : hom A B),
    monotone (hom B C) (hom A C) (fun f => comp f g);
  comp_cont_l : forall {A B C} (f : hom B C),
    continuous (Build_mono_fun _ (comp_mono_l f));
  comp_cont_r : forall {A B C} (g : hom A B),
    continuous (Build_mono_fun _ (comp_mono_r g));
  comp_id_l : forall {A B} (f : hom A B),
    comp (id_mor B) f = f;
  comp_id_r : forall {A B} (f : hom A B),
    comp f (id_mor A) = f;
  comp_assoc : forall {A B C D}
    (h : hom C D) (g : hom B C) (f : hom A B),
    comp (comp h g) f = comp h (comp g f);
}.
\end{lstlisting}

Two aspects of this mixin deserve comment.

\paragraph{Separate continuity (DD-002).}
\Cref{def:cpo-enriched-cat} asks that composition be
Scott-continuous in each argument separately; by
\cref{prop:joint-continuity} (Abramsky--Jung Lemma~3.2.6), this is
equivalent to joint continuity of the binary operation.  The separate
form is chosen because the joint form would require the product CPO
$\texttt{hom}\;A\;B \times \texttt{hom}\;B\;C$,  which is defined in
\texttt{Products.v}.  Importing it at the structures stage would create
a dependency cycle:
\[
  \texttt{Enriched.v} \to \texttt{Products.v} \to \texttt{CPOTheory.v}
    \to \texttt{Enriched.v}.
\]
Separate continuity breaks the cycle entirely.
The joint form is derived as a theorem in \texttt{EnrichedTheory.v}
once \texttt{Products.v} is available.

\paragraph{Monotonicity before continuity (DD-003).}
The \texttt{comp\_mono\_l}/\texttt{comp\_mono\_r} fields appear before
\texttt{comp\_cont\_l}/\texttt{comp\_cont\_r} because the
\texttt{continuous} predicate takes a \texttt{mono\_fun} argument;
placing the monotonicity witnesses first ensures they are in scope when
the continuity fields are elaborated.

\paragraph{Self-enrichment.}
The canonical example is \textbf{CPO} itself: taking
$\texttt{hom}\;D\;E = [D \to_c E]$ (the function-space CPO of
\cref{prop:function-space}), $\texttt{id\_mor} = \texttt{cont\_id}$,
and $\texttt{comp} = \texttt{cont\_comp}$, the separate-continuity
conditions reduce to the fact that post- and pre-composition with a
fixed continuous function are themselves continuous.  This instance
is registered in \texttt{Function.v}.

% ·· Locally continuous functors ····························
\subsection{Locally Continuous Functors}
\label{sec:meth-lc-functors}

% Source: theory/EnrichedTheory.v (~706 lines)

A locally continuous endofunctor (\cref{def:lc-functor}) acts
continuously on each hom-CPO.  While the HB hierarchy declares
\texttt{IsLocallyContinuous} as a mixin (bundled into
\texttt{LocallyContinuousFunctor}), the main theory uses a
\emph{plain record} instead:

\begin{lstlisting}[
  caption={The \texttt{lc\_functor} record
    (\texttt{EnrichedTheory.v}).  A plain Rocq record rather than an HB
    structure; see text for the rationale.},
  label={lst:lcfunctor}
]
Record lc_functor (C : CPOEnrichedCat.type) := {
  lcf_obj  : C -> C;
  lcf_mor  : forall (A B : C),
    hom A B -> hom (lcf_obj A) (lcf_obj B);
  lcf_mono : forall (A B : C),
    monotone _ _ (lcf_mor A B);
  lcf_cont : forall (A B : C),
    continuous (Build_mono_fun _ (lcf_mono A B));
  lcf_id   : forall (A : C),
    lcf_mor A A (id_mor A) = id_mor (lcf_obj A);
  lcf_comp : forall (A B D : C)
    (f : hom B D) (g : hom A B),
    lcf_mor A D (comp f g) =
      comp (lcf_mor B D f) (lcf_mor A B g);
}.
\end{lstlisting}

The decision to use a plain record rather than an HB structure (DD-018)
is driven by two constraints.  First, registering HB instances for
functors on \texttt{CPOEnrichedCat.type} triggers universe
inconsistencies when natural transformations quantify over all objects
in the enriched category.  Second, identity, composition, and
whiskering of functors produce new \texttt{lc\_functor} values; with HB
structures each composition would require a new instance registration,
whereas plain records compose directly.  A conversion function
\texttt{lc\_functor\_of\_hb} bridges the two representations;
the rationale and the decision not to use universe polymorphism are
discussed in \cref{sec:meth-design-decisions} (DD-019).

Identity and composition of locally continuous functors are defined on
\texttt{lc\_functor}: \texttt{id\_lc\_functor} packages the identity
endofunctor, and \texttt{comp\_lc\_functor} composes two LC functors,
with the expected associativity and unit laws.

\paragraph{Joint continuity of composition.}
A key derived result is that composition---axiomatised as separately
continuous---is jointly continuous on the product hom-CPO.  The proof
proceeds in two stages.  Stage~A works in a product-free setting: given
chains $(g_n)$ in $\texttt{hom}\;B\;D$ and $(f_n)$ in
$\texttt{hom}\;A\;B$, the diagonal chain $n \mapsto g_n \circ f_n$
satisfies $\bigsqcup_n (g_n \circ f_n) = (\bigsqcup_n g_n)
\circ (\bigsqcup_n f_n)$, proved via the separate-continuity rewrite
lemmas and the $\max(n,m)$ diagonal bound.  Stage~B wraps this as a
\texttt{cont\_fun} on the product hom-CPO using the product CPO
projections.  The two-stage split is necessary because a direct proof
over the product type causes irreconcilable HB coercion failures
between \texttt{CPO.sort} and \texttt{CPOEnrichedCat.sort}.

\paragraph{HB coercion workarounds (DD-016).}
The coercion difficulties in the joint-continuity proof are symptomatic
of a broader issue: HB-synthesised coercion paths
(e.g.\ \texttt{CPOEnrichedCat.sort} $\to$ \texttt{CPO.sort}) resolve
ambiguously when the elaborator encounters intermediate terms with
under-determined types.  \texttt{EnrichedTheory.v} employs five
systematic workarounds:
\begin{itemize}
  \item \texttt{etransitivity} instead of \texttt{le\_trans with X},
    deferring the intermediate term to a unification variable;
  \item explicit \texttt{@}-qualified projections
    (e.g.\ \texttt{@comp\_assoc C A B C' D});
  \item \texttt{unshelve econstructor} for one-field-at-a-time record
    construction;
  \item \texttt{Arguments \{C\}} directives to keep the enriched
    category implicit; and
  \item \texttt{cbn [mf\_fun]} followed by \texttt{cbv beta} to expose
    term structure hidden by nested coercions before \texttt{rewrite}.
\end{itemize}
These workarounds are required only when compiling from source
(\texttt{dune build}); pre-compiled \texttt{.vo} files resolve the
coercions correctly.  The broader implications of these friction
points are discussed in \cref{ch:discussion}.

% ·· Natural transformations ································
\subsection{Natural Transformations}
\label{sec:meth-nat-trans}

% Source: theory/NatTrans.v (~518 lines)

Enriched natural transformations (\cref{def:enriched-nat-trans}) between
locally continuous functors are formalised as a plain record
\texttt{nat\_trans~F~G} over two \texttt{lc\_functor} values, carrying
a family of components $\alpha_A \in \texttt{hom}\;(F\,A)\;(G\,A)$ and
a naturality proof.  As with \texttt{lc\_functor}, the plain-record
approach (DD-018) avoids universe conflicts and simplifies composition.

\texttt{NatTrans.v} establishes the algebraic structure of
natural transformations:
\begin{itemize}
  \item \textbf{Identity and vertical composition.}
    The identity natural transformation has components
    $(\mathrm{id}_F)_A = \texttt{id\_mor}\;(F\,A)$; the vertical
    composite $(\beta \cdot \alpha)_A = \beta_A \circ \alpha_A$ is
    constructed by pasting naturality squares
    (\cref{prop:vcomp}).  Associativity and unit laws follow from the
    categorical laws of the base enriched category.
  \item \textbf{Horizontal composition (whiskering).}
    Left whiskering $(H * \alpha)_A = H_{FA,GA}(\alpha_A)$ applies a
    functor to each component; right whiskering
    $(\alpha * H)_A = \alpha_{HA}$ reindexes along a functor.
    The interchange law
    $H * (\beta \cdot \alpha) = (H * \beta) \cdot (H * \alpha)$
    follows from the functoriality of~$H$.
  \item \textbf{Pointwise CPO structure.}
    The pointwise order
    $\alpha \sqsubseteq \beta \iff \forall A.\;
    \alpha_A \sqsubseteq \beta_A$ makes the collection of
    natural transformations $F \Rightarrow G$ into an $\omega$-CPO
    (\cref{prop:nat-trans-cpo}).
    The supremum $(\bigsqcup_n \alpha^{(n)})_A =
    \bigsqcup_n \alpha^{(n)}_A$ satisfies naturality by the
    three-step argument: expand via left separate continuity, apply
    componentwise naturality, and collapse via right separate
    continuity.  Antisymmetry of the pointwise order uses
    \texttt{FunctionalExtensionality} and
    \texttt{ProofIrrelevance}.
\end{itemize}

% ·· Domain equations ·······································
\subsection{Domain Equations}
\label{sec:meth-domain-equations}

% Source: theory/DomainEquations.v (~446 lines)

Recursive domain equations---type-level fixed points of the form
$D \cong F(D, D)$ for a mixed-variance bifunctor~$F$---require
all of the enriched categorical machinery developed so far.
\Cref{sec:bg-bilimits} presents the mathematical theory:
\cref{def:mixed-bifunctor} defines mixed-variance bifunctors,
\cref{def:bilimit} defines bilimits, and
\cref{thm:smyth-plotkin} guarantees their existence.
Here we describe the formalisation.

\paragraph{The mixed-variance bifunctor mixin.}
The data of a mixed-variance bifunctor is recorded in
\texttt{HasMixedEndo} (\texttt{Enriched.v}): an object action
\texttt{MF\_obj} and a morphism action \texttt{MF\_mor} that is
contravariant in its first argument and covariant in its second.
The axioms---identity, composition, separate monotonicity, and
separate continuity---are deferred to a second mixin
\texttt{IsMixedLocallyContinuous} in \texttt{DomainEquations.v}
(DD-005), following the same cycle-avoidance strategy as DD-002.

From the six separate axioms, the derived properties are straightforward:
joint monotonicity of \texttt{MF\_mor}
and the joint-supremum identity
$F(\bigsqcup_n f_n,\; \bigsqcup_n g_n) = \bigsqcup_n F(f_n, g_n)$,
proved via the same $\max(n,m)$ diagonal argument used for composition
in \cref{sec:meth-lc-functors}.

\paragraph{EP-pair lifting.}
Given an ep-pair $(e, p) : D \to E$
(\cref{def:ep-pair}), the bifunctor lifts it to an ep-pair
$F(D,D) \to F(E,E)$ with embedding $F(p, e)$ and projection $F(e, p)$.
The contravariant argument receives the projection (reversing
direction), and the covariant argument receives the embedding.  The
retraction and deflation laws follow from the functoriality
and monotonicity axioms.

\paragraph{The approximation sequence.}
Starting from an initial object~$D_0$ and an ep-pair
$D_0 \hookrightarrow F(D_0, D_0)$, repeated application of the
bifunctor builds an expanding sequence
\[
  D_0 \xrightarrow{e_0} F(D_0, D_0) \xrightarrow{e_1}
    F^2(D_0, D_0) \xrightarrow{e_2} \cdots
\]
where each step is an ep-pair.  This is packaged as an
\texttt{ep\_chain} record.

\paragraph{The bilimit and its existence (DD-017).}
The \texttt{BilimitData} record mechanises \cref{def:bilimit}: it
specifies a bilimit object~$D_\infty$;
cone ep-pairs $D_n \hookrightarrow D_\infty$ for each approximant
(\cref{def:ep-cone}),
with compatibility conditions ensuring the cone commutes with the
approximation sequence; a proof that
$\mathrm{id}_{D_\infty} = \bigsqcup_n (e^\infty_n \circ p^\infty_n)$
(the bilimit condition \eqref{eq:bilimit-identity}); and the
ROLL/UNROLL isomorphism of \cref{thm:smyth-plotkin}.

The existence of such a bilimit is asserted as an axiom:
\begin{quote}
\texttt{Axiom bilimit\_exists :}\\
\quad\texttt{forall \{C : MixedLCFunctor.type\} (D0 : C)}\\
\quad\quad\texttt{(ep0 : ep\_pair D0 (MF\_obj D0 D0)),}\\
\quad\quad\texttt{BilimitData D0 ep0.}
\end{quote}
Constructing the bilimit requires the $\omega$-product CPO
$\prod_n D_n$---the product of countably many CPOs---which is not yet
in the library.  Rather than axiomatise the bilimit's
\emph{properties} ad hoc, the library axiomatises its
\emph{existence} via a fully specified record; all consequences (the
ROLL/UNROLL isomorphism, the deflation chain, and the identity
$\texttt{id}_{D_\infty} = \bigsqcup_n (e_n \circ p_n)$) are then proved
from the record fields with no further axioms.  A worked
example---the recursive domain $D \cong [D \to_c D]_\bot$---is
presented in \cref{sec:res-domain-equations}.

% ── PCF formalization ─────────────────────────────────────────────
\section{PCF Formalization}
\label{sec:meth-pcf}

The PCF$_v$ formalisation is inherited from the original
Benton--Kennedy--Varming development~\cite{BKV2009} and updated to
the current Rocq/HB infrastructure; the differences are catalogued in
\cref{sec:disc-bkv}.  It exercises the library's CPO constructions
(products, lifting, function spaces, the Kleene fixed-point operator)
in a concrete setting: three files totalling 2303 lines implement
syntax (\texttt{PCF\_Syntax.v}, 804 lines), big-step evaluation
(\texttt{PCF\_Operational.v}, 332 lines), and denotational semantics
(\texttt{PCF\_Denotational.v}, 1167 lines).
This section describes the key design choices that distinguish the
mechanised development from the mathematical presentation in
\cref{sec:bg-denotational}; the reader should refer to that section
for the definitions of types, terms, evaluation rules, and the
denotational interpretation.

% ·· Syntax and typing ······································
\subsection{Syntax and Typing}
\label{sec:meth-pcf-syntax}

% Source: lang/PCF_Syntax.v (804 lines)

\paragraph{Intrinsic typing and ANF (DD-009).}
Following Benton, Kennedy, and Varming~\cite{BKV2009}, all syntactic
objects are \emph{intrinsically typed}: \texttt{Value~:~Env~$\to$~Ty~$\to$~Type}
and \texttt{Exp~:~Env~$\to$~Ty~$\to$~Type} are mutually inductive
families indexed by a typing environment $\Gamma$ and a type $\tau$.
Every constructor preserves the typing invariant, so there are no
separate typing judgements, no well-formedness side-conditions, and no
need to prove that two typing derivations of the same term are equal.
Variables are typed de~Bruijn indices (\texttt{Var~$\Gamma$~$\tau$}),
with constructors \texttt{ZVAR} (index~$0$) and \texttt{SVAR}
(successor index).  The term grammar is in administrative normal form
(\cref{def:pcfv-terms}): function arguments must be values, and
intermediate computations are sequenced by \texttt{LET}.

Compared with the original Benton--Kennedy names, we drop the
\texttt{T} prefix (since all terms are typed by construction) and rename
\texttt{TINT} to \texttt{NLIT} and \texttt{TBOOL} to \texttt{BLIT}
to reflect the underlying Rocq types; a complete naming comparison
appears in \cref{sec:disc-bkv}.

\paragraph{The \texttt{var\_case} combinator (DD-013).}
Several definitions need to case-split a variable
\texttt{Var~($\tau$~::~$\Gamma$)~$\sigma$} into the head position
(\texttt{ZVAR}, where $\sigma = \tau$) and the tail (\texttt{SVAR~x},
where $x$~:~\texttt{Var~$\Gamma$~$\sigma$}).  A na\"ive dependent
\texttt{match} introduces the opaque axiom \texttt{JMeq\_eq} from
\texttt{Program.Equality}, which blocks kernel reduction and causes
\texttt{reflexivity} to fail on computation rules.  The
\texttt{var\_case} combinator avoids this by using a dependent match
whose return type is itself a function:

\begin{lstlisting}[
  caption={The \texttt{var\_case} combinator (\texttt{PCF\_Syntax.v}).
    The return clause produces a \emph{function} of \texttt{hz} and
    \texttt{hs}, so that type equalities are resolved by
    $\iota$-reduction rather than equality transport.},
  label={lst:varcase}
]
Definition var_case {Γ τ} {P : Ty → Type}
    (hz : P τ) (hs : forall σ, Var Γ σ → P σ)
    : forall σ, Var (τ :: Γ) σ → P σ :=
  fun σ (x : Var (τ :: Γ) σ) =>
    match x in Var ctx σ' return
          match ctx with
          | []        => P σ'
          | τ₀ :: Γ₀ =>
              P τ₀ → (forall σ, Var Γ₀ σ → P σ) → P σ'
          end
    with
    | ZVAR   => fun z _  => z
    | SVAR y => fun _ s  => s _ y
    end hz hs.
\end{lstlisting}

The key insight is that the match returns a \emph{function} that
receives \texttt{hz} and \texttt{hs} as arguments after the match,
sidestepping the need for an equality cast inside either branch.
As a result, \texttt{var\_case~hz~hs~ZVAR} reduces to \texttt{hz}
and \texttt{var\_case~hz~hs~(SVAR~y)} reduces to \texttt{hs~\_~y}
by pure $\iota$+$\beta$-reduction---no axioms involved.

This combinator is used in four places: \texttt{ren\_ext} (renaming
extension), \texttt{subst\_ext} (substitution extension),
\texttt{single\_subst}, and \texttt{double\_subst}.  In each case,
the computation rules hold \emph{definitionally}---for example,
\texttt{single\_subst~v~$\tau$~ZVAR~$\equiv$~v}---which is
essential for the soundness proof, where evaluation rules must
compose cleanly with substitution lemmas.

\paragraph{Renaming bootstrap (DD-009).}
Substitutions (\texttt{Subst~$\Gamma$~$\Gamma'$} $=$
$\forall\,\tau$, \texttt{Var~$\Gamma$~$\tau$} $\to$
\texttt{Value~$\Gamma'$~$\tau$}$\,$) require weakening: extending
a substitution under a binder must shift each value by one de~Bruijn
level.  Following McBride's technique, renamings
(\texttt{Ren~$\Gamma$~$\Gamma'$} $=$
$\forall\,\tau$, \texttt{Var~$\Gamma$~$\tau$} $\to$
\texttt{Var~$\Gamma'$~$\tau$}$\,$) are defined first, and the
weakening operation \texttt{wkVal} is defined by applying the
canonical renaming \texttt{ren\_wk} that shifts every index by one.
This avoids a complex dependent shift operator on values.

% ·· Operational semantics ··································
\subsection{Operational Semantics}
\label{sec:meth-pcf-operational}

% Source: lang/PCF_Operational.v (332 lines)

\paragraph{Big-step CBV evaluation (DD-010).}
The operational semantics is a big-step evaluation relation
\texttt{Eval~:~CExp~$\tau$~$\to$~CValue~$\tau$~$\to$~Prop}
(notation $e \Downarrow v$) on closed terms, where
\texttt{CExp~$\tau$} $=$ \texttt{Exp~[]~$\tau$} and
\texttt{CValue~$\tau$} $=$ \texttt{Value~[]~$\tau$}.
The nine evaluation rules (\cref{sec:bg-pcf-syntax}) are
transcribed directly; the two rules that involve substitution use the
operations from \texttt{PCF\_Syntax.v}:
\begin{itemize}
  \item \textsc{E-Let}: evaluates $e_1 \Downarrow v_1$, then
    \texttt{singleSubstExp~$v_1$~$e_2$}~$\Downarrow v_2$.
  \item \textsc{E-App}: for
    $\texttt{APP}~(\texttt{FIX}~\tau_1~\tau_2~\mathit{body})~v_{\mathrm{arg}}$,
    evaluates
    \texttt{doubleSubstExp~$v_{\mathrm{arg}}$~(FIX~$\tau_1$~$\tau_2$~$\mathit{body}$)~$\mathit{body}$},
    which substitutes the argument for index~$0$ and the recursive
    self-reference for index~$1$.
\end{itemize}

Big-step evaluation is chosen over small-step reduction because the
soundness theorem takes the natural form
$e \Downarrow v \implies \llbracket e \rrbracket = \eta(\llbracket v \rrbracket)$
(\cref{thm:soundness}), matching the structure of the denotational
semantics directly.  With intrinsic typing, there are no stuck
states: every closed value is already in normal form (the only closed
value of function type is \texttt{FIX}), so no
progress or preservation theorems are needed.

The file also derives \texttt{eval\_deterministic}: if
$e \Downarrow v_1$ and $e \Downarrow v_2$ then $v_1 = v_2$, proved
by double induction on the two derivations.

% ·· Denotational semantics ·································
\subsection{Denotational Semantics}
\label{sec:meth-pcf-denotational}

% Source: lang/PCF_Denotational.v (1167 lines)

The denotational semantics maps types to CPOs, values to continuous
maps, and expressions to continuous maps into the lifted codomain,
following the mathematical presentation of \cref{sec:bg-pcf-denotational}.
The Rocq formalisation differs from the on-paper version in three
significant ways: the denotation functions are point-free, the renaming
lemma bootstraps the substitution lemma, and the \texttt{FIX} case
uses a single combinator application.

\paragraph{Point-free mutual fixpoint (DD-011).}
Two approaches were considered for packaging the denotation:

\begin{enumerate}
  \item \textbf{Three-pass}: define raw functions
    \texttt{sem\_val\_fn}/\texttt{sem\_exp\_fn}, prove monotonicity,
    prove continuity, then package as \texttt{cont\_fun}.  Every
    syntactic case must be handled three times.
  \item \textbf{Single-pass}: each case is a point-free pipeline of
    library combinators (\texttt{cont\_comp}, \texttt{cont\_pair},
    \texttt{cont\_curry}, \texttt{FIXP}, \texttt{kleislir},
    \texttt{ret}, etc.) that are already packaged as
    \texttt{cont\_fun}.  The mutual fixpoint directly produces
    \texttt{cont\_fun} values with no extra proof obligations.
\end{enumerate}

The single-pass approach was chosen.  For example:
\begin{itemize}
  \item \texttt{VAR~x} denotes \texttt{sem\_var~x}, the iterated
    projection $\pi_2^k \circ \pi_1$ corresponding to the de~Bruijn
    index.
  \item \texttt{LET~$e_1$~$e_2$} denotes
    \texttt{kleislir(sem\_exp~$e_2$~$\circ$~swap)~$\circ$~$\langle$id,
    sem\_exp~$e_1$$\rangle$}: pair the environment with the result
    of~$e_1$ (which may be $\bot$), then thread through the Kleisli
    extension of~$e_2$.
  \item \texttt{FIX~$\tau_1$~$\tau_2$~$\mathit{body}$} denotes
    \texttt{FIXP~$\circ$~flip(curry(flip(curry(sem\_exp~$\mathit{body}$))))},
    applying the continuous fixed-point operator
    (\cref{thm:kleene}) after currying the body's two-argument
    interpretation.
\end{itemize}

The downside of the point-free style is that \texttt{simpl} unfolds
all combinator internals, producing goals with deeply nested
projections and coercions.  The proofs in \texttt{PCF\_Denotational.v}
use \texttt{change} rather than \texttt{simpl} to control unfolding,
and selectively make definitions \texttt{Opaque} during the
renaming and substitution mutual proofs.

\paragraph{Renaming route to the substitution lemma (DD-012).}
The substitution lemma
($\llbracket e[\sigma] \rrbracket\;\gamma =
 \llbracket e \rrbracket\;(\llbracket \sigma \rrbracket\;\gamma)$)
is the critical interface between syntax and semantics: it is the tool
that allows the soundness proof to reduce the \textsc{E-Let} and
\textsc{E-App} cases to the induction hypotheses.  The proof follows
the bootstrapping strategy of \cref{eq:sem-subst-ext}, translated into
the Rocq setting as a chain of mutual lemmas:
\[
  \texttt{sem\_ren\_ext}
    \;\to\; \texttt{sem\_val\_ren}
    \;\to\; \texttt{sem\_ren\_wk}
    \;\to\; \texttt{sem\_subst\_ext}
    \;\to\; \texttt{sem\_val\_subst}.
\]
The renaming lemma (\texttt{sem\_val\_ren}/\texttt{sem\_exp\_ren})
is proved first by mutual induction on terms; it is needed to
establish that syntactic weakening (\texttt{wkVal~v}) corresponds to
dropping the outermost environment component
($\llbracket \texttt{wkVal}\;v \rrbracket\;(d,\gamma)
= \llbracket v \rrbracket\;\gamma$).  This \texttt{sem\_ren\_wk}
lemma is then used to prove the substitution extension lemma
\texttt{sem\_subst\_ext}, which in turn powers the binder cases
(\texttt{FIX}, \texttt{LET}) of the full substitution lemma.

The end-user interface consists of two corollaries:
\begin{itemize}
  \item \texttt{sem\_single\_subst}: relates \texttt{singleSubstExp} to
    extending the semantic environment by one value.
  \item \texttt{sem\_double\_subst}: relates \texttt{doubleSubstExp} to
    extending by two values (argument and self-reference).
\end{itemize}
These are the only substitution lemmas needed by the soundness and
adequacy proofs in \cref{ch:results}.

\medskip

With the syntax, operational semantics, and denotational semantics in
place, the remaining task is to show that the two agree.
Soundness~(\cref{sec:res-soundness}) proceeds by structural induction
on the big-step evaluation derivation, following DD-014; each rule
reduces to an application of the substitution corollaries above.
Adequacy~(\cref{sec:res-adequacy}) is more involved: it constructs a
type-indexed logical relation between domain elements and closed terms,
and the \texttt{FIX} case appeals to \texttt{fixp\_ind}
(\cref{thm:kleene}) to establish the admissibility of the
relation---the strategy introduced by Pitts and inherited from
Benton, Kennedy, and Varming (DD-015).
